---
title: Ferrari bloodlines
description: An early experiment in reading Ferrari history as families, generations and interrupted lines.
date: 2000-01-01
date-modified: 2026-08-15
categories: [dont-forget]
draft: true
echo: fenced
---

<script>document.body.style.cssText += "--post-accent:#c41230;--post-end-ink:#fff8f6;--post-end-hover:#ffffff"</script>

```{r}
#| label: setup
#| include: false

source(here::here("blog", "_R", "post_setup.R"))

install_missing_packages(c(
  "tidyverse",
  "highcharter",
  "gh",
  "httr2"
))
```

Ferrari history is easy to present as a catalogue: one beautiful car after another. It is harder—and more interesting—to ask which cars belong to the same idea.

This is an early workspace for that question. The data comes from [`motorsport-legends`](https://github.com/jbkunst/motorsport-legends), where production periods, technical attributes and relationships are being reviewed. This post consumes its output; it does not redefine the history locally.

## Read the data

```{r}
#| label: load-ferrari-data
#| message: false

library(tidyverse)
library(highcharter)

ferrari_data_url <- paste0(
  "https://api.github.com/repos/jbkunst/motorsport-legends/contents/",
  "data/cars/ferrari_past_models_enriched.csv"
)

ferrari_response <- httr2::request(ferrari_data_url) |>
  httr2::req_headers(
    Accept = "application/vnd.github.raw+json",
    Authorization = paste("Bearer", gh::gh_token()),
    `X-GitHub-Api-Version` = "2022-11-28"
  ) |>
  httr2::req_perform()

ferrari <- ferrari_response |>
  httr2::resp_body_string() |>
  I() |>
  readr::read_csv(show_col_types = FALSE)

ferrari
```

The snapshot currently contains `r nrow(ferrari)` models arranged into `r n_distinct(ferrari$main_family)` principal families. It already has the structure of a genealogy, although its historical coverage is not finished.

```{r}
#| label: summarise-data-coverage

coverage <- ferrari |>
  summarise(
    models = n(),
    production_period = sum(!is.na(end_year)),
    start_year_only = sum(is.na(end_year)),
    lineage = sum(!is.na(lineage)),
    predecessor = sum(!is.na(predecessor))
  ) |>
  pivot_longer(everything(), names_to = "field", values_to = "models") |>
  mutate(
    field = recode(
      field,
      production_period = "Known production period",
      start_year_only = "Start year only",
      lineage = "Assigned lineage",
      predecessor = "Known predecessor",
      .default = "All models"
    )
  )

coverage
```

That distinction matters visually. A long bar should mean that we know a production period, not that an algorithm invented one. Models whose end date is still unknown remain points.

## Families before chronology

```{r}
#| label: count-models-by-family

family_summary <- ferrari |>
  count(main_family, sort = TRUE) |>
  mutate(main_family = fct_reorder(main_family, n))

ggplot(family_summary, aes(n, main_family)) +
  geom_segment(
    aes(x = 0, xend = n, yend = main_family),
    linewidth = 1.4,
    colour = "#d9d2c3"
  ) +
  geom_point(size = 3.2, colour = "#c41230") +
  labs(
    title = "Eleven ways through Ferrari history",
    subtitle = "The current taxonomy is useful, but some families still contain very different stories.",
    x = "Models",
    y = NULL
  )
```

The largest lane, sports-racing prototypes, is likely too broad for the final story. Road-car lineages can be read as generations; competition history often branches by regulation, programme and derived model instead.

## A first timeline

The official Ferrari archive gives each car room to be seen before it is classified. This experiment follows that rhythm: the timeline provides the structure, while one selected model occupies the foreground.

```{r}
#| label: show-featured-ferrari
#| echo: false
#| results: asis

featured <- ferrari |>
  filter(model == "365 GTB4") |>
  slice(1)

escape_html <- function(x, attribute = FALSE) {
  htmltools::htmlEscape(coalesce(as.character(x), ""), attribute = attribute)
}

cat(
  '<style>',
  '.ferrari-feature{display:grid;grid-template-columns:minmax(250px,.78fr) minmax(0,1.22fr);min-height:390px;background:#161412;color:#fff;margin:2rem 0 1rem;overflow:hidden}',
  '.ferrari-feature-copy{display:flex;flex-direction:column;justify-content:flex-end;padding:clamp(2rem,5vw,4.5rem)}',
  '.ferrari-feature-kicker{color:#ef2438;font-size:.72rem;letter-spacing:.18em;text-transform:uppercase;margin-bottom:1rem}',
  '.ferrari-feature h3{color:#fff;font-size:clamp(2.3rem,5vw,4.8rem);font-weight:200;line-height:.95;margin:0 0 1rem}',
  '.ferrari-feature-meta{color:#d6d0c9;font-size:.88rem;letter-spacing:.06em;text-transform:uppercase}',
  '.ferrari-feature-description{color:#bcb5ad;max-width:34rem;margin:1.3rem 0 0}',
  '.ferrari-feature-media{position:relative;min-height:390px;background:#2b2824}',
  '.ferrari-feature-media img{width:100%;height:100%;object-fit:cover;display:block}',
  '@media(max-width:760px){.ferrari-feature{grid-template-columns:1fr}.ferrari-feature-media{min-height:260px;order:-1}.ferrari-feature-copy{padding:2rem}}',
  '</style>',
  '<div class="ferrari-feature">',
  '<div class="ferrari-feature-copy">',
  '<div class="ferrari-feature-kicker">Selected bloodline</div>',
  sprintf('<h3 id="ferrari-feature-name">%s</h3>', escape_html(featured$model)),
  sprintf(
    '<div id="ferrari-feature-meta" class="ferrari-feature-meta">%s · %s</div>',
    escape_html(featured$start_year),
    escape_html(featured$main_family)
  ),
  sprintf(
    '<p id="ferrari-feature-description" class="ferrari-feature-description">%s</p>',
    escape_html(featured$description)
  ),
  '</div>',
  '<div class="ferrari-feature-media">',
  sprintf(
    '<img id="ferrari-feature-image" src="%s" alt="%s">',
    escape_html(featured$image_url, TRUE),
    escape_html(featured$model, TRUE)
  ),
  '</div>',
  '</div>',
  sep = "\n"
)
```

Click a model to bring its photograph and description into the foreground.

```{r}
#| label: prepare-ferrari-timeline
#| include: false

family_levels <- ferrari |>
  count(main_family, sort = TRUE) |>
  pull(main_family) |>
  rev()

timeline <- ferrari |>
  mutate(
    family_id = match(main_family, family_levels) - 1L,
    start_date = as.Date(sprintf("%s-01-01", start_year)),
    end_date = as.Date(if_else(
      is.na(end_year),
      NA_character_,
      sprintf("%s-12-31", end_year)
    )),
    relationship = coalesce(predecessor_relationship, "origin or unclassified"),
    lineage_label = coalesce(lineage, "Lineage not assigned")
  )

period_points <- timeline |>
  filter(!is.na(end_year)) |>
  transmute(
    x = datetime_to_timestamp(start_date),
    x2 = datetime_to_timestamp(end_date),
    y = family_id,
    name = model,
    color = "#c41230",
    custom = pmap(
      list(
        family = main_family,
        lineage = lineage_label,
        start = start_year,
        end = end_year,
        engine = engine_architecture,
        layout = engine_location,
        relationship = relationship,
        predecessor = coalesce(predecessor, "—"),
        image = coalesce(image_url, ""),
        description = coalesce(description, "")
      ),
      list
    )
  ) |>
  list_parse()

date_points <- timeline |>
  filter(is.na(end_year)) |>
  transmute(
    x = datetime_to_timestamp(start_date),
    y = family_id,
    name = model,
    custom = pmap(
      list(
        family = main_family,
        lineage = lineage_label,
        start = start_year,
        engine = engine_architecture,
        layout = engine_location,
        relationship = relationship,
        predecessor = coalesce(predecessor, "—"),
        image = coalesce(image_url, ""),
        description = coalesce(description, "")
      ),
      list
    )
  ) |>
  list_parse()
```

```{r}
#| code-fold: true
#| label: show-ferrari-timeline
#| column: screen-inset

tooltip_html <- JS(
  "function () {",
  "  const c = this.point.custom || {};",
  "  const end = c.end ? '–' + c.end : ' · end year pending';",
  "  const image = c.image ? '<img src=\"' + c.image + '\" alt=\"\" style=\"display:block;width:260px;height:146px;object-fit:cover;margin:-12px -14px 12px\">' : '';",
  "  return '<div style=\"padding:12px 14px;min-width:220px\">' +",
  "    image +",
  "    '<div style=\"font-size:16px;font-weight:600;color:#28120f\">' + this.point.name + '</div>' +",
  "    '<div style=\"margin:2px 0 10px;color:#8f1d23\">' + c.start + end + '</div>' +",
  "    '<div><b>Family</b> · ' + c.family + '</div>' +",
  "    '<div><b>Lineage</b> · ' + c.lineage + '</div>' +",
  "    '<div><b>Engine</b> · ' + (c.engine || 'pending') + ' · ' + (c.layout || 'pending') + '</div>' +",
  "    '<div><b>Previous</b> · ' + c.predecessor + '</div>' +",
  "    '<div><b>Relation</b> · ' + c.relationship.replaceAll('_', ' ') + '</div>' +",
  "  '</div>';",
  "}"
)

select_model_js <- JS(
  "function () {",
  "  const c = this.custom || {};",
  "  const image = document.getElementById('ferrari-feature-image');",
  "  const name = document.getElementById('ferrari-feature-name');",
  "  const meta = document.getElementById('ferrari-feature-meta');",
  "  const description = document.getElementById('ferrari-feature-description');",
  "  if (name) name.textContent = this.name;",
  "  if (meta) meta.textContent = c.start + (c.end ? '–' + c.end : '') + ' · ' + c.family;",
  "  if (description) description.textContent = c.description || c.lineage;",
  "  if (image && c.image) { image.src = c.image; image.alt = this.name; }",
  "}"
)

highchart() |>
  hc_add_dependency("modules/xrange.js") |>
  hc_chart(
    backgroundColor = "#f5f1e8",
    height = 720,
    zoomType = "x",
    spacing = c(28, 24, 22, 12),
    style = list(fontFamily = "IBM Plex Sans, sans-serif")
  ) |>
  hc_title(
    text = "Ferrari bloodlines",
    align = "left",
    style = list(color = "#28120f", fontSize = "28px", fontWeight = "300")
  ) |>
  hc_subtitle(
    text = "Production periods are bars. Unfinished research remains visible as points.",
    align = "left",
    style = list(color = "#765f56", fontSize = "14px")
  ) |>
  hc_xAxis(
    type = "datetime",
    min = datetime_to_timestamp(as.Date("1947-01-01")),
    gridLineWidth = 1,
    gridLineColor = "#ded7ca",
    lineColor = "#9b8e80",
    tickColor = "#9b8e80"
  ) |>
  hc_yAxis(
    categories = family_levels,
    title = list(text = NULL),
    gridLineColor = "#ded7ca",
    labels = list(style = list(color = "#4a3731", fontSize = "12px"))
  ) |>
  hc_plotOptions(
    series = list(
      cursor = "pointer",
      point = list(events = list(click = select_model_js))
    )
  ) |>
  hc_add_series(
    type = "xrange",
    name = "Known production period",
    data = period_points,
    borderColor = "#f5f1e8",
    borderWidth = 1.2,
    borderRadius = 1,
    pointWidth = 10,
    states = list(
      hover = list(
        brightness = 0.08,
        borderColor = "#ffffff",
        borderWidth = 2
      )
    )
  ) |>
  hc_add_series(
    type = "scatter",
    name = "End year pending",
    data = date_points,
    color = "#c41230",
    marker = list(
      radius = 4,
      symbol = "circle",
      lineWidth = 1.5,
      lineColor = "#c41230",
      fillColor = "#f5f1e8"
    )
  ) |>
  hc_tooltip(
    useHTML = TRUE,
    shared = FALSE,
    followPointer = TRUE,
    backgroundColor = "#fffdf8",
    borderColor = "#c9bdae",
    borderRadius = 8,
    shadow = TRUE,
    formatter = tooltip_html
  ) |>
  hc_legend(
    align = "center",
    verticalAlign = "bottom",
    itemStyle = list(color = "#4a3731", fontWeight = "400")
  ) |>
  hc_credits(
    enabled = TRUE,
    text = "Data: motorsport-legends · work in progress",
    href = "https://github.com/jbkunst/motorsport-legends"
  )
```

This is not yet the final genealogy. The visual language is the useful part of the experiment: warm paper rather than dashboard white, Ferrari red used sparingly, restrained typography, and uncertainty shown as part of the data.

The final chart should make a selected family rise above the catalogue. Direct succession can remain continuous; conceptual inheritance should cross a gap with a quieter connector. Competition derivatives may need their own layer instead of occupying the same visual grammar as road-car generations.

## What comes next

The historical work belongs in `motorsport-legends`. Once a family has been reviewed there, this post should only need a refreshed CSV to tell a more complete story.
