---
title: Decomposing vintage risk into age, cohort and period
description: Reconstruct monthly loan hazards from public Bondora loans and split vintage risk into age, cohort and period effects.
author: Joshua Kunst
date: 2026-09-10
categories: [R, credit risk, vintage analysis, APC]
lang: en
draft: true
toc: true
---

```{r}
#| label: setup
#| include: false

source("../../_R/post_setup.R")

install_missing_packages(c(
  "dplyr",
  "ggplot2",
  "lubridate",
  "openxlsx",
  "purrr",
  "scales",
  "tibble",
  "tidyr"
))
```

Credit vintages are a common way to follow how default changes after loans are
originated. The [Federal Reserve](https://www.federalreserve.gov/bankinforeg/bcreg20130819a1.pdf)
describes vintage loss models, also called **age-cohort-time models**, for retail
credit portfolios. The [Banco de la República of Colombia](https://www.banrep.gov.co/en/publications-research/financial-stability-report/box-1-second-half-2024)
uses an **age-period-cohort decomposition** directly on credit vintages to help
understand where changes in credit risk come from.

The idea is simple. Every observation in a vintage has three dates attached to it:

- **Age**: how many months old the loan is.
- **Cohort**: the month when the loan was originated.
- **Period**: the calendar month when we observe it.

For monthly data they are linked exactly:

\[
\text{period} = \text{cohort} + \text{age}.
\]

The easiest way to understand the decomposition is to build the data from the
bottom up. We will start with one loan, reconstruct its monthly history, repeat the
same operation for every loan, summarise those rows into monthly default hazards,
and only then separate the result into AGE, COHORT and PERIOD.

## Public loan data

[Go & Grow's public statistics page](https://goandgrow.eu/en/public-statistics/)
links a downloadable Bondora loan dataset. The file contains one row per loan and
includes origination, default, contract-end and reporting dates. That is enough to
reconstruct the monthly history needed here.

The source is a live snapshot, so results can move when the provider refreshes the
file. `DefaultDate` is Bondora's own default event, so throughout this post
**default means the event recorded in that field**.

```{r}
#| label: download-public-loans

library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)

data_url <- "https://sabanners001.blob.core.windows.net/statistics/public/loan_dataset_investor.xlsx"

# Keep the large source file outside the repository and reuse it on later renders.
cache_dir <- tools::R_user_dir("jkunst-blog", which = "cache")
data_file <- file.path(cache_dir, "bondora-loan-dataset.xlsx")
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)

if (!file.exists(data_file)) {
  download.file(data_url, data_file, mode = "wb")
}
```

The public workbook is wide, but only a few fields are needed here. Reading just
those columns keeps the example smaller even if the source grows.

```{r}
#| label: read-loan-fields

needed_columns <- c(
  "ReportAsOfEOD",
  "LoanId",
  "LoanDate",
  "Country",
  "Status",
  "DefaultDate",
  "ContractEndDate",
  "LastPaymentOn"
)

# Read only the header first so column positions do not have to be hard-coded.
header <- openxlsx::read.xlsx(
  data_file,
  rows = 1,
  colNames = FALSE,
  skipEmptyCols = FALSE
) |>
  unlist(use.names = FALSE) |>
  as.character()

column_index <- match(needed_columns, header)
missing_columns <- needed_columns[is.na(column_index)]

if (length(missing_columns) > 0) {
  stop(
    "Bondora file layout changed. Missing columns: ",
    paste(missing_columns, collapse = ", ")
  )
}

loans_raw <- openxlsx::read.xlsx(
  data_file,
  cols = column_index,
  colNames = TRUE,
  detectDates = TRUE,
  check.names = FALSE
)
```

For a compact example I keep Estonian originations from 2022 through 2025 and at
most the first 36 months of performance. The recent cohorts are useful because
they make the incomplete vintage triangle visible.

```{r}
#| label: prepare-loans

country_code <- "EE"
cohort_from <- as.Date("2022-01-01")
cohort_to <- as.Date("2025-12-01")
max_age <- 36L

# Excel exports can store dates as date cells, timestamps or serial numbers.
as_bondora_date <- function(x) {
  if (inherits(x, "Date")) return(x)
  if (inherits(x, "POSIXt")) return(as.Date(x))
  if (is.numeric(x)) return(openxlsx::convertToDate(x))
  as.Date(substr(as.character(x), 1, 10))
}

loans <- loans_raw |>
  transmute(
    loan_id = LoanId,
    country = Country,
    status = Status,
    loan_date = as_bondora_date(LoanDate),
    default_date = as_bondora_date(DefaultDate),
    contract_end_date = as_bondora_date(ContractEndDate),
    last_payment_date = as_bondora_date(LastPaymentOn),
    report_date = as_bondora_date(ReportAsOfEOD)
  ) |>
  mutate(
    cohort = floor_date(loan_date, unit = "month"),
    default_month = floor_date(default_date, unit = "month"),
    contract_end_month = floor_date(contract_end_date, unit = "month"),
    last_payment_month = floor_date(last_payment_date, unit = "month"),
    report_month = floor_date(report_date, unit = "month")
  ) |>
  filter(
    country == country_code,
    cohort >= cohort_from,
    cohort <= cohort_to
  ) |>
  mutate(
    # Default ends the history. Otherwise we stop at repayment or the reporting month.
    exit_month = case_when(
      !is.na(default_month) ~ default_month,
      status == "Repaid" & !is.na(last_payment_month) ~ last_payment_month,
      status == "Repaid" & !is.na(contract_end_month) ~ contract_end_month,
      TRUE ~ report_month
    ),
    exit_age = interval(cohort, exit_month) %/% months(1),
    default_age = if_else(
      !is.na(default_month),
      interval(cohort, default_month) %/% months(1),
      NA_real_
    ),
    observed_age = pmin(exit_age, max_age)
  ) |>
  filter(
    observed_age >= 1,
    is.na(default_age) | default_age >= 1
  )

loans |>
  summarise(
    loans = n(),
    defaults = sum(!is.na(default_age) & default_age <= max_age),
    first_cohort = min(cohort),
    last_cohort = max(cohort)
  )
```

At this point there is still one row per loan. We know when the loan started and,
if applicable, when default occurred, but there is not yet one row for every month
that the loan was observed.

## One loan, month by month

Start with one loan that eventually defaulted. Looking at one case makes the
expansion easy to understand before doing the same thing thousands of times.

```{r}
#| label: choose-one-loan

one_loan <- loans |>
  filter(!is.na(default_age), default_age <= max_age) |>
  arrange(abs(default_age - 6)) |>
  slice(1) |>
  select(loan_id, cohort, observed_age, default_age)

one_loan
```

`uncount()` turns that one record into one row per observed month. Age 1 is the
first month after origination, so its period is the cohort month plus one month.
The default indicator is one only in the month when default first appears.

```{r}
#| label: expand-one-loan

one_loan_months <- one_loan |>
  uncount(observed_age, .id = "age") |>
  mutate(
    period = cohort %m+% months(age),
    default = as.integer(!is.na(default_age) & age == default_age)
  ) |>
  select(loan_id, cohort, age, period, default)

one_loan_months
```

A loan defaulting at age 6 appears in six monthly rows, but contributes exactly
**one** new default: at age 6. We do not create rows after that default.

## The same transformation for every loan

Now put the same logic in a small function. It takes one loan and returns its
monthly rows.

```{r}
#| label: define-expand-loan

expand_loan <- function(loan_id, cohort, observed_age, default_age) {
  tibble::tibble(
    loan_id = loan_id,
    cohort = cohort,
    age = seq_len(as.integer(observed_age))
  ) |>
    mutate(
      period = cohort %m+% months(age),
      default = as.integer(!is.na(default_age) & age == default_age)
    )
}
```

`pmap()` applies the function to every loan. Each loan returns a small monthly
table and `list_rbind()` joins all of them.

```{r}
#| label: expand-all-loans

loan_months <- loans |>
  select(loan_id, cohort, observed_age, default_age) |>
  purrr::pmap(expand_loan) |>
  purrr::list_rbind()

loan_months |>
  arrange(loan_id, age) |>
  slice_head(n = 12)
```

We have moved from **one row per loan** to:

`loan_id × cohort × age × period × default`.

## From monthly records to vintage hazards

Now group the loan rows by cohort, age and period. For each combination we count
how many loans are still exposed and how many new defaults happen in that month.

```{r}
#| label: build-vintage-long

vintage_long <- loan_months |>
  summarise(
    exposure = n(),
    defaults = sum(default),
    .by = c(cohort, age, period)
  ) |>
  mutate(hazard = defaults / exposure) |>
  arrange(cohort, age)

vintage_long |>
  slice_head(n = 15)
```

For each row:

\[
\text{hazard} =
\frac{\text{new defaults during the month}}
     {\text{loans exposed at the start of the month}}.
\]

So the hazard is simply the **monthly default rate among loans that reached that
month without defaulting before**. It is not an accumulated default rate.

The long table is the object we will use for the rest of the analysis. Each row has
three identities at once. For example, one hazard can belong to age 4, the January
2023 cohort and May 2023 period. APC will ask whether that value looks high because
of the age, the cohort or the period.

## The familiar vintage triangle

A vintage table is often displayed with cohort down the rows and age across the
columns. That shape is useful for looking at the data, so take only the latest
twelve cohorts and the first twelve months.

```{r}
#| label: show-vintage-triangle

triangle_cohorts <- vintage_long |>
  distinct(cohort) |>
  arrange(desc(cohort)) |>
  slice_head(n = 12) |>
  pull(cohort)

vintage_triangle <- vintage_long |>
  filter(
    cohort %in% triangle_cohorts,
    age <= 12
  ) |>
  transmute(
    cohort = format(cohort, "%Y-%m"),
    age,
    hazard = scales::percent(hazard, accuracy = 0.01)
  ) |>
  pivot_wider(
    names_from = age,
    names_prefix = "M",
    values_from = hazard
  ) |>
  arrange(cohort)

vintage_triangle
```

The empty cells for the newest cohorts simply mean those loans have not reached
those ages yet.

Most importantly, **the triangle is only a view**. We do not use this wide table
for the decomposition. The source remains `vintage_long`, where cohort, age,
period, exposure, defaults and hazard are explicit columns.

## Raw age profile

Before separating the three parts, group the long table only by age. This answers
a simple question: among all loans that reached each age, how often did a new
default happen during that month?

```{r}
#| label: plot-raw-age-hazard

age_hazard <- vintage_long |>
  summarise(
    exposure = sum(exposure),
    defaults = sum(defaults),
    .by = age
  ) |>
  mutate(hazard = defaults / exposure)

ggplot(age_hazard, aes(age, hazard)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1)) +
  labs(
    title = "Default hazard by loan age",
    subtitle = "Raw age profile before separating cohort and period",
    x = "Age (months)",
    y = "Monthly default hazard"
  )
```

This curve is useful, but it still mixes the three parts. Age 6 for an old cohort
happened in a different period from age 6 for a newer cohort. A particularly risky
cohort can also make some ages look worse simply because many of its loans are
present there.

## Sequential APC decomposition

We now return to `vintage_long`. The triangle was only for looking at the data; the
decomposition uses the complete long table.

The hazard is first adjusted slightly so rows with zero defaults can still be
transformed. Then we use a logit transformation because it lets us add the three
effects on one scale.

\[
q = \frac{defaults + 0.5}{exposure + 1}.
\]

We then split the result into:

\[
y \approx \mu + AGE + COHORT + PERIOD.
\]

The simple version used here does that in order. First AGE is estimated, then
COHORT from what remains, and finally PERIOD from what remains after both. Exposure
is used as the weight throughout.

```{r}
#| label: estimate-sequential-apc

alpha <- 0.5
beta <- 1

apc_cells <- vintage_long |>
  mutate(
    q = (defaults + alpha) / (exposure + beta),
    y_logit = qlogis(q)
  )

mu <- weighted.mean(apc_cells$y_logit, apc_cells$exposure)

age_effects <- apc_cells |>
  mutate(residual = y_logit - mu) |>
  summarise(
    effect = weighted.mean(residual, exposure),
    weight = sum(exposure),
    .by = age
  ) |>
  mutate(effect = effect - weighted.mean(effect, weight))

cells_after_age <- apc_cells |>
  left_join(
    age_effects |>
      select(age, age_effect = effect),
    by = "age"
  )

cohort_effects <- cells_after_age |>
  mutate(residual = y_logit - mu - age_effect) |>
  summarise(
    effect = weighted.mean(residual, exposure),
    weight = sum(exposure),
    .by = cohort
  ) |>
  mutate(effect = effect - weighted.mean(effect, weight))

cells_after_cohort <- cells_after_age |>
  left_join(
    cohort_effects |>
      select(cohort, cohort_effect = effect),
    by = "cohort"
  )

period_effects <- cells_after_cohort |>
  mutate(residual = y_logit - mu - age_effect - cohort_effect) |>
  summarise(
    effect = weighted.mean(residual, exposure),
    weight = sum(exposure),
    .by = period
  ) |>
  mutate(effect = effect - weighted.mean(effect, weight))

apc_decomposition <- cells_after_cohort |>
  left_join(
    period_effects |>
      select(period, period_effect = effect),
    by = "period"
  ) |>
  mutate(
    fitted_logit = mu + age_effect + cohort_effect + period_effect,
    fitted_hazard = plogis(fitted_logit)
  )
```

The result is now much easier to inspect: one AGE effect for every month of age,
one COHORT effect for every origination month, and one PERIOD effect for every
calendar month. The plots below put each effect back on a hazard scale. The dashed
line is the overall baseline.

```{r}
#| label: plot-age-effect

baseline_hazard <- plogis(mu)

ggplot(age_effects, aes(age, plogis(mu + effect))) +
  geom_hline(yintercept = baseline_hazard, linetype = "dashed") +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1)) +
  labs(
    title = "AGE effect",
    subtitle = "Which loan ages show more risk?",
    x = "Age (months)",
    y = "Default hazard"
  )
```

```{r}
#| label: plot-cohort-effect
#| column: page

ggplot(cohort_effects, aes(cohort, plogis(mu + effect))) +
  geom_hline(yintercept = baseline_hazard, linetype = "dashed") +
  geom_line() +
  geom_point() +
  scale_x_date(date_breaks = "6 months", date_labels = "%Y-%m") +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1)) +
  labs(
    title = "COHORT effect",
    subtitle = "Which origination cohorts show more risk after AGE is removed?",
    x = "Origination cohort",
    y = "Default hazard"
  )
```

```{r}
#| label: plot-period-effect
#| column: page

ggplot(period_effects, aes(period, plogis(mu + effect))) +
  geom_hline(yintercept = baseline_hazard, linetype = "dashed") +
  geom_line() +
  geom_point() +
  scale_x_date(date_breaks = "6 months", date_labels = "%Y-%m") +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1)) +
  labs(
    title = "PERIOD effect",
    subtitle = "Which calendar periods show more risk after AGE and COHORT are removed?",
    x = "Calendar period",
    y = "Default hazard"
  )
```

The interpretation is direct. A peak in AGE points to ages where default is more
likely. A peak in COHORT points to origination months that look riskier. A peak in
PERIOD points to calendar months where the loans observed at that time look
riskier.

## One important caveat

This post uses a **sequential** decomposition because it is easy to see step by
step. But AGE, COHORT and PERIOD are linked by `period = cohort + age`, so there is
no unique way to separate all three without making an additional choice. Here the
choice is the order: AGE first, then COHORT, then PERIOD.

That is why these plots should be read as one transparent way of dividing the
observed vintage risk, not as the only possible decomposition. A later comparison
can use a simultaneous APC method and check whether the same risky ages, cohorts
and periods appear.

## References

Federal Reserve Board (2013), *Capital Planning at Large Bank Holding Companies*: vintage loss models, also called age-cohort-time models, for retail credit portfolios.  
Banco de la República (2024), *Decomposition of the Performance of Credit Vintages*: an age-period-cohort application to credit vintages in its Financial Stability Report.
