---
title: Decomposing vintage risk into age, cohort and period
description: Reconstruct monthly loan hazards from public Bondora loans and split vintage risk into age, cohort and period effects.
author: Joshua Kunst
date: 2026-09-10
categories: [R, credit risk, vintage analysis, APC]
lang: en
draft: true
toc: true
---

```{r}
#| label: setup
#| include: false

source("blog/_R/post_setup.R")

install_missing_packages(c(
  "dplyr",
  "ggplot2",
  "knitr",
  "lubridate",
  "openxlsx",
  "purrr",
  "scales",
  "tibble",
  "tidyr"
))
```

Credit vintages are a common way to follow how default changes after loans are
originated. The [Federal Reserve](https://www.federalreserve.gov/bankinforeg/bcreg20130819a1.pdf)
describes vintage loss models, also called **age-cohort-time models**, for retail
credit portfolios. The [Banco de la República of Colombia](https://www.banrep.gov.co/en/publications-research/financial-stability-report/box-1-second-half-2024)
uses an **age-period-cohort decomposition** directly on credit vintages to help
understand where changes in credit risk come from.

The idea is simple. Every observation in a vintage has three dates attached to it:

- **Age**: how many months old the loan is.
- **Cohort**: the month when the loan was originated.
- **Period**: the calendar month when we observe it.

For monthly data they are linked exactly:

$$
\text{period} = \text{cohort} + \text{age}.
$$

We will build the data from the bottom up: one loan, its monthly history, all loans,
a long vintage table, a familiar vintage triangle, and finally the APC decomposition.

## Public loan data

[Go & Grow's public statistics page](https://goandgrow.eu/en/public-statistics/)
links a downloadable Bondora loan dataset. The workbook contains one row per loan
and a second sheet with a data dictionary.

```{r}
#| label: download-public-loans

library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)

data_url <- "https://sabanners001.blob.core.windows.net/statistics/public/loan_dataset_investor.xlsx"

# Keep the large source file outside the repository and reuse it on later renders.
cache_dir <- tools::R_user_dir("jkunst-blog", which = "cache")
data_file <- file.path(cache_dir, "bondora-loan-dataset.xlsx")
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)

if (!file.exists(data_file)) {
  download.file(data_url, data_file, mode = "wb")
}
```

We only need six fields from the 31 columns in the loan sheet. The names below
come directly from the current public workbook.

```{r}
#| label: read-loan-fields

needed_columns <- c(
  "loan_id",
  "country",
  "loan_issued_at",
  "loan_status",
  "is_default",
  "months_on_book"
)

# Read row 1 first so we can select only the columns needed from the large workbook.
header <- openxlsx::read.xlsx(
  data_file,
  sheet = "Loan Dataset",
  rows = 1,
  colNames = FALSE,
  skipEmptyCols = FALSE
) |>
  unlist(use.names = FALSE) |>
  as.character()

column_index <- match(needed_columns, header)

if (anyNA(column_index)) stop("Missing Bondora fields: ", paste(needed_columns[is.na(column_index)], collapse = ", "))

loans_raw <- openxlsx::read.xlsx(
  data_file,
  sheet = "Loan Dataset",
  cols = column_index,
  detectDates = TRUE,
  check.names = FALSE
) |>
  tibble::as_tibble()

loans_raw
```

Printing the tibble is intentional: it shows the number of rows and columns, the
fields we loaded and a few real observations before we transform anything.

The data contains loans from several countries. Before choosing one for the
example, look at their sizes.

```{r}
#| label: count-countries

country_counts <- loans_raw |>
  count(country, sort = TRUE)

country_counts
```

For the decomposition we use Estonia. It is not the largest country in the current
file, but it has substantial and continuous monthly originations throughout the
2018--2022 window. That makes it a cleaner example than markets with long gaps in
origination, while keeping country differences out of the AGE, COHORT and PERIOD
effects.

The workbook also explains the fields we use. In particular, `months_on_book` is
the number of months a loan has been in active status.

```{r}
#| label: inspect-data-dictionary

loan_dictionary <- openxlsx::read.xlsx(
  data_file,
  sheet = "Dataset Dictionary"
) |>
  tibble::as_tibble() |>
  filter(Column %in% needed_columns) |>
  select(Column, Description)

knitr::kable(
  loan_dictionary,
  col.names = c("Field", "Description"),
  align = c("l", "l")
)
```

For this example we keep Estonian cohorts from 2018 through 2022 and follow at
most the first 36 months. Starting before 2020 is useful because the same calendar
period can meet different cohorts at very different ages.

```{r}
#| label: prepare-loans

country_name <- "Estonia"
cohort_from <- as.Date("2018-01-01")
cohort_to <- as.Date("2022-12-01")
max_age <- 36L

loans <- loans_raw |>
  mutate(
    # Excel stores this field as a serial date in the current workbook.
    loan_date = as.Date(loan_issued_at, origin = "1899-12-30"),
    cohort = floor_date(loan_date, unit = "month"),
    months_on_book = as.integer(months_on_book),
    is_default = as.logical(is_default),
    # Use Bondora's MOB convention directly: a defaulted loan defaults at this age.
    default_age = if_else(is_default, months_on_book, NA_integer_),
    observed_age = pmin(months_on_book, max_age)
  ) |>
  filter(
    country == country_name,
    cohort >= cohort_from,
    cohort <= cohort_to,
    observed_age >= 1
  ) |>
  select(
    loan_id,
    country,
    loan_status,
    cohort,
    months_on_book,
    is_default,
    default_age,
    observed_age
  )

loans
```

At this point there is still one row per loan. We know its cohort, how many months
we observe it, and whether default happened at the end of that observed path.

## One loan, month by month

Start with one loan that defaulted around age 6.

```{r}
#| label: choose-one-loan

one_loan <- loans |>
  filter(is_default, default_age <= max_age) |>
  arrange(abs(default_age - 6)) |>
  slice(1) |>
  select(loan_id, cohort, observed_age, default_age)

one_loan
```

`uncount()` turns that one record into one row per observed month. Age 1 is the
first month after origination. The default flag is one only in the month when the
default first appears.

```{r}
#| label: expand-one-loan

one_loan_months <- one_loan |>
  uncount(observed_age, .id = "age") |>
  mutate(
    period = cohort %m+% months(age),
    default = as.integer(!is.na(default_age) & age == default_age)
  ) |>
  select(loan_id, cohort, age, period, default)

one_loan_months
```

A loan defaulting at age 6 appears in six monthly rows but contributes exactly one
new default: at age 6. We do not create rows after that default.

## The same idea for more loans

Now put the same logic in a small function.

```{r}
#| label: define-expand-loan

expand_loan <- function(loan_id, cohort, observed_age, default_age) {
  tibble::tibble(
    loan_id = loan_id,
    cohort = cohort,
    age = seq_len(as.integer(observed_age))
  ) |>
    mutate(
      period = cohort %m+% months(age),
      default = as.integer(!is.na(default_age) & age == default_age)
    )
}
```

`pmap()` makes the row-by-row idea explicit. Here we use only three loans so the
result stays small enough to inspect.

```{r}
#| label: expand-three-loans

three_loans <- loans |>
  slice_head(n = 3) |>
  select(loan_id, cohort, observed_age, default_age)

three_loan_months <- three_loans |>
  purrr::pmap(expand_loan) |>
  purrr::list_rbind()

three_loan_months
```

For the full portfolio we do not need to build thousands of tiny tibbles. The same
operation can be written directly with `uncount()`, which is simpler and faster.

```{r}
#| label: expand-all-loans

loan_months <- loans |>
  select(loan_id, cohort, observed_age, default_age) |>
  uncount(observed_age, .id = "age") |>
  mutate(
    period = cohort %m+% months(age),
    default = as.integer(!is.na(default_age) & age == default_age)
  ) |>
  select(loan_id, cohort, age, period, default)

loan_months
```

We have moved from one row per loan to:

`loan_id × cohort × age × period × default`.

## From monthly records to vintage hazards

Now group those rows by cohort, age and period. For each combination we count how
many loans reached the month and how many new defaults happened there.

```{r}
#| label: build-vintage-long

vintage_long <- loan_months |>
  summarise(
    exposure = n(),
    defaults = sum(default),
    .by = c(cohort, age, period)
  ) |>
  mutate(hazard = defaults / exposure) |>
  arrange(cohort, age)

vintage_long
```

For each row:

$$
\text{hazard} =
\frac{\text{new defaults during the month}}
     {\text{loans exposed at the start of the month}}.
$$

So the hazard is simply the monthly default rate among loans that reached that
month without defaulting before. It is not an accumulated default rate.

Each row now has three identities at once. A hazard can be high because of the age
of the loans, because that cohort is different, because something happened in that
calendar period, or because of a mixture of those things.

## The familiar vintage triangle

The same long table can be displayed with cohort down the rows and age across the
columns. Here we take twelve cohorts around 2020 and their first twelve months.

```{r}
#| label: show-vintage-triangle
#| column: page

triangle_cohorts <- seq.Date(
  as.Date("2019-06-01"),
  as.Date("2020-05-01"),
  by = "month"
)

vintage_triangle <- vintage_long |>
  filter(
    cohort %in% triangle_cohorts,
    age <= 12
  ) |>
  select(cohort, age, hazard) |>
  mutate(
    cohort = format(cohort, "%Y-%m"),
    hazard = scales::percent(hazard, accuracy = 0.01)
  ) |>
  pivot_wider(
    names_from = age,
    names_prefix = "M",
    values_from = hazard
  ) |>
  arrange(cohort)

vintage_triangle
```

The triangle is only another view of `vintage_long`; it is not a different data
source. A calendar month cuts across the triangle diagonally because different
cohorts reach the same period at different ages.

For example, look at April 2020 directly in the long table:

```{r}
#| label: inspect-one-period

period_example <- as.Date("2020-04-01")

vintage_long |>
  filter(
    period == period_example,
    cohort >= as.Date("2018-01-01"),
    cohort <= as.Date("2020-03-01")
  ) |>
  select(cohort, age, period, exposure, defaults, hazard) |>
  arrange(cohort)
```

It is the **same period**, but the 2018 loans are much older than the 2019 loans,
and the early-2020 loans are only a few months old. A raw vintage view cannot tell
us whether a difference comes from age, cohort or the common calendar period.

That is why including pre-2020 cohorts is useful here. If the PERIOD component
later changes around 2020, it is a pattern worth inspecting. The decomposition by
itself does not prove that COVID caused the change.

## Cumulative default by cohort

Before decomposing anything, look at a classic vintage view: cumulative default as
loans get older. For a cleaner overview, we pool the twelve monthly cohorts into
origination years from 2018 to 2022.

```{r}
#| label: plot-cumulative-vintages
#| column: page

annual_vintages <- loan_months |>
  mutate(cohort_year = year(cohort)) |>
  summarise(
    exposure = n(),
    defaults = sum(default),
    .by = c(cohort_year, age)
  ) |>
  arrange(cohort_year, age) |>
  group_by(cohort_year) |>
  mutate(
    initial_loans = first(exposure),
    cumulative_defaults = cumsum(defaults),
    cumulative_default_rate = cumulative_defaults / initial_loans
  ) |>
  ungroup()

p_vintages <- annual_vintages |>
  mutate(cohort_year = factor(cohort_year)) |>
  ggplot(aes(age, cumulative_default_rate, colour = cohort_year, group = cohort_year)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
  labs(
    title = "Cumulative default by vintage",
    subtitle = "Estonian loans originated from 2018 to 2022",
    x = "Age (months)",
    y = "Cumulative default rate",
    colour = "Cohort year"
  )

p_vintages
```

Different lines can have different shapes for many reasons. The loans were
originated in different years, they reached each age in different calendar
periods, and they passed through events such as 2020 at different points in their
life. That is exactly the mixture we want to separate.

The yearly grouping above is only for a cleaner descriptive chart. The APC
decomposition below still uses the original monthly cohorts in `vintage_long`.
This raw vintage chart is also a natural candidate for the post preview.

## Raw age profile

We can also ignore cohort and period for a moment and group only by age.

```{r}
#| label: plot-raw-age-hazard

age_hazard <- vintage_long |>
  summarise(
    exposure = sum(exposure),
    defaults = sum(defaults),
    .by = age
  ) |>
  mutate(hazard = defaults / exposure)

age_hazard

ggplot(age_hazard, aes(age, hazard)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1)) +
  labs(
    title = "Default hazard by loan age",
    subtitle = "Raw age profile before separating cohort and period",
    x = "Age (months)",
    y = "Monthly default hazard"
  )
```

This curve is useful, but it still mixes the three parts. Age 12 for a 2018 cohort
happened in a different period from age 12 for a 2020 cohort.

## Sequential APC decomposition

We now return to `vintage_long`. The triangle and cumulative curves were only for
looking at the data; the decomposition uses the complete long table.

Some cohort-age-period rows have no defaults. Their raw hazard is exactly zero,
and `logit(0)` is not finite. To avoid that, use a small **add-half correction**:
add half a default and half a non-default to every row.

$$
q = \frac{defaults + 0.5}{exposure + 1}.
$$

The `0.5` in the numerator is the half default. The denominator increases by `1`
because we added `0.5` to each of the two outcomes: default and no default. In
large rows the change is tiny; its purpose is to keep the logit finite.

Why introduce another variable, `y`, instead of decomposing the default rate
directly? A probability is bounded between zero and one, so additive effects are
awkward on that scale. The logit maps the adjusted hazard `q` to the whole real
line:

$$
y = \operatorname{logit}(q) = \log\left(\frac{q}{1-q}\right).
$$

On this scale the components can be added. After combining them, `plogis()` is the
inverse transformation that returns us to a default probability:

$$
q = \operatorname{plogis}(y) = \frac{1}{1 + e^{-y}}.
$$

We then split `y` into a mean plus AGE, COHORT and PERIOD effects:

$$
y \approx \mu + AGE + COHORT + PERIOD.
$$

The simple version used here does that in order. First AGE is estimated, then
COHORT from what remains, and finally PERIOD from what remains after both. Exposure
is used as the weight throughout.

```{r}
#| label: estimate-sequential-apc

pseudo_count <- 0.5

apc_cells <- vintage_long |>
  mutate(
    q = (defaults + pseudo_count) / (exposure + 2 * pseudo_count),
    y_logit = qlogis(q)
  )

mu <- weighted.mean(apc_cells$y_logit, apc_cells$exposure)

age_effects <- apc_cells |>
  mutate(residual = y_logit - mu) |>
  summarise(
    effect = weighted.mean(residual, exposure),
    weight = sum(exposure),
    .by = age
  ) |>
  mutate(effect = effect - weighted.mean(effect, weight))

cells_after_age <- apc_cells |>
  left_join(
    age_effects |>
      select(age, age_effect = effect),
    by = "age"
  )

cohort_effects <- cells_after_age |>
  mutate(residual = y_logit - mu - age_effect) |>
  summarise(
    effect = weighted.mean(residual, exposure),
    weight = sum(exposure),
    .by = cohort
  ) |>
  mutate(effect = effect - weighted.mean(effect, weight))

cells_after_cohort <- cells_after_age |>
  left_join(
    cohort_effects |>
      select(cohort, cohort_effect = effect),
    by = "cohort"
  )

period_effects <- cells_after_cohort |>
  mutate(residual = y_logit - mu - age_effect - cohort_effect) |>
  summarise(
    effect = weighted.mean(residual, exposure),
    weight = sum(exposure),
    .by = period
  ) |>
  mutate(effect = effect - weighted.mean(effect, weight))

apc_decomposition <- cells_after_cohort |>
  left_join(
    period_effects |>
      select(period, period_effect = effect),
    by = "period"
  ) |>
  mutate(
    mean_effect = mu,
    fitted_logit = mean_effect + age_effect + cohort_effect + period_effect,
    residual = y_logit - fitted_logit,
    fitted_hazard = plogis(fitted_logit),
    reconstructed_logit = fitted_logit + residual,
    reconstructed_q = plogis(reconstructed_logit)
  ) |>
  select(
    cohort,
    age,
    period,
    exposure,
    defaults,
    q,
    y_logit,
    mean_effect,
    age_effect,
    cohort_effect,
    period_effect,
    residual,
    fitted_hazard,
    reconstructed_q
  )

apc_decomposition
```

There **is** a residual. The three effects explain the systematic part allocated by
this sequential procedure, while each row can still differ from that fitted value.
The exact row-level identity is:

$$
y = \mu + AGE + COHORT + PERIOD + residual.
$$

The table above lets us see every piece together. We can also check the sum
directly on a few rows:

```{r}
#| label: check-apc-sum

apc_decomposition |>
  mutate(
    component_sum = mean_effect + age_effect + cohort_effect + period_effect + residual,
    difference = y_logit - component_sum
  ) |>
  select(
    cohort,
    age,
    period,
    y_logit,
    mean_effect,
    age_effect,
    cohort_effect,
    period_effect,
    residual,
    component_sum,
    difference
  ) |>
  slice_head(n = 10)
```

`difference` should be zero apart from numerical precision. On the probability
scale, adding the residual back also recovers `q`.

```{r}
#| label: check-apc-reconstruction

apc_decomposition |>
  summarise(
    max_rate_difference = max(abs(q - reconstructed_q))
  )
```

The three component plots below stay on the scale where the decomposition is
additive, but exponentiate each effect so it is easier to read. `1x` means no
change in odds, `1.5x` means 50% higher odds and `0.5x` means half the odds. These
are **relative effects**, not three separate default rates.

```{r}
#| label: plot-age-effect

ggplot(age_effects, aes(age, exp(effect))) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.01, suffix = "x")) +
  labs(
    title = "AGE effect",
    subtitle = "1x means average odds; above 1x means higher odds",
    x = "Age (months)",
    y = "Odds multiplier"
  )
```

```{r}
#| label: plot-cohort-effect
#| column: page

ggplot(cohort_effects, aes(cohort, exp(effect))) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_line() +
  geom_point() +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.01, suffix = "x")) +
  labs(
    title = "COHORT effect",
    subtitle = "Relative odds after AGE is removed",
    x = "Origination cohort",
    y = "Odds multiplier"
  )
```

```{r}
#| label: plot-period-effect
#| column: page

ggplot(period_effects, aes(period, exp(effect))) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = as.Date("2020-03-01"), linetype = "dotted") +
  geom_line() +
  geom_point() +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.01, suffix = "x")) +
  labs(
    title = "PERIOD effect",
    subtitle = "Relative odds after AGE and COHORT are removed",
    x = "Calendar period",
    y = "Odds multiplier"
  )
```

The three lines show the components separately. To see them working together, use
the same vintage geometry again. Cohort and age define the position of every tile,
and `period = cohort + age` is therefore implicit in that position. The fill is the
full fitted hazard after adding AGE, COHORT and PERIOD in logit space and applying
`plogis()`.

```{r}
#| label: plot-fitted-apc-surface
#| column: page

ggplot(apc_decomposition, aes(age, cohort, fill = fitted_hazard)) +
  geom_tile() +
  scale_x_continuous(
    breaks = seq(0, max_age, by = 6),
    expand = expansion(mult = 0)
  ) +
  scale_y_date(
    date_breaks = "1 year",
    date_labels = "%Y",
    expand = expansion(mult = 0)
  ) +
  scale_fill_viridis_c(labels = scales::label_percent(accuracy = 0.1)) +
  labs(
    title = "Fitted APC hazard surface",
    subtitle = "Vertical patterns = AGE, horizontal = COHORT, diagonal = PERIOD",
    x = "Age (months)",
    y = "Origination cohort",
    fill = "Fitted hazard"
  )
```

This tile plot is also a useful visual check on the decomposition. Vertical bands
come from AGE because age is fixed down a column. Horizontal bands come from
COHORT because cohort is fixed across a row. PERIOD moves diagonally through the
matrix because the same calendar month is reached by older loans from earlier
cohorts and younger loans from later cohorts.

The interpretation of the component plots is therefore relative: a peak in AGE
means higher odds at those ages, a peak in COHORT means higher odds for those
origination months after AGE is removed, and a peak in PERIOD means higher odds in
those calendar months after AGE and COHORT are removed. The fitted tile is where
those pieces are combined back into a default hazard.

## One important caveat

This post uses a **sequential** decomposition because it is easy to see step by
step. But AGE, COHORT and PERIOD are linked by `period = cohort + age`, so there is
no unique way to separate all three without making an additional choice. Here the
choice is the order: AGE first, then COHORT, then PERIOD.

That is why these plots should be read as one transparent way of dividing the
observed vintage risk, not as the only possible decomposition. A later comparison
can use a simultaneous APC method and check whether the same risky ages, cohorts
and periods appear.

## References

Federal Reserve Board (2013), *Capital Planning at Large Bank Holding Companies*: vintage loss models, also called age-cohort-time models, for retail credit portfolios.  
Banco de la República (2024), *Decomposition of the Performance of Credit Vintages*: an age-period-cohort application to credit vintages in its Financial Stability Report.