---
title: Decomposing vintage risk into age, cohort and period
description: Reconstruct monthly loan hazards from public Bondora loans and split vintage risk into age, cohort and period effects.
author: Joshua Kunst
date: 2026-09-10
categories: [R, credit risk, vintage analysis, APC]
lang: en
draft: true
toc: true
---

```{r}
#| label: setup
#| include: false

source("../../_R/post_setup.R")

install_missing_packages(c(
  "dplyr",
  "ggplot2",
  "lubridate",
  "openxlsx",
  "purrr",
  "scales",
  "tibble",
  "tidyr"
))
```

Credit vintages are a common way to follow how default changes after loans are
originated. The [Federal Reserve](https://www.federalreserve.gov/bankinforeg/bcreg20130819a1.pdf)
describes vintage loss models, also called **age-cohort-time models**, for retail
credit portfolios. The [Banco de la República of Colombia](https://www.banrep.gov.co/en/publications-research/financial-stability-report/box-1-second-half-2024)
uses an **age-period-cohort decomposition** directly on credit vintages to help
understand where changes in credit risk come from.

The idea is simple. Every observation in a vintage has three dates attached to it:

- **Age**: how many months old the loan is.
- **Cohort**: the month when the loan was originated.
- **Period**: the calendar month when we observe it.

For monthly data they are linked exactly:

\[
\text{period} = \text{cohort} + \text{age}.
\]

The easiest way to understand the decomposition is to build the data from the
bottom up. We will start with one loan, reconstruct its monthly history, repeat the
same operation for every loan, summarise those rows into monthly default hazards,
and only then separate the result into AGE, COHORT and PERIOD.

## Public loan data

[Go & Grow's public statistics page](https://goandgrow.eu/en/public-statistics/)
links a downloadable Bondora loan dataset. The file contains one row per loan and
includes origination, default, contract-end and reporting dates. That is enough to
reconstruct the monthly history needed here.

The source is a live snapshot, so results can move when the provider refreshes the
file. `DefaultDate` is Bondora's own default event, so throughout this post
**default means the event recorded in that field**.

```{r}
#| label: download-public-loans

library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)

data_url <- "https://sabanners001.blob.core.windows.net/statistics/public/loan_dataset_investor.xlsx"

# Keep the large source file outside the repository and reuse it on later renders.
cache_dir <- tools::R_user_dir("jkunst-blog", which = "cache")
data_file <- file.path(cache_dir, "bondora-loan-dataset.xlsx")
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)

if (!file.exists(data_file)) {
  download.file(data_url, data_file, mode = "wb")
}
```

The public workbook is wide, but only a few fields are needed here.

```{r}
#| label: read-loan-fields

needed_columns <- c(
  "ReportAsOfEOD",
  "LoanId",
  "LoanDate",
  "Country",
  "Status",
  "DefaultDate",
  "ContractEndDate",
  "LastPaymentOn"
)

# Read the header first so column positions do not have to be hard-coded.
header <- openxlsx::read.xlsx(
  data_file,
  rows = 1,
  colNames = FALSE,
  skipEmptyCols = FALSE
) |>
  unlist(use.names = FALSE) |>
  as.character()

column_index <- match(needed_columns, header)
missing_columns <- needed_columns[is.na(column_index)]

if (length(missing_columns) > 0) {
  stop(
    "Bondora file layout changed. Missing columns: ",
    paste(missing_columns, collapse = ", ")
  )
}

loans_raw <- openxlsx::read.xlsx(
  data_file,
  cols = column_index,
  colNames = TRUE,
  detectDates = TRUE,
  check.names = FALSE
) |>
  tibble::as_tibble()

loans_raw
```

Printing each main tibble is intentional. It lets us see the number of rows and
columns, the fields and a few values before moving to the next step.

For the example I keep Estonian cohorts from 2018 through 2022 and at most the
first 36 months of each loan. Starting before 2020 is useful: the same calendar
shock can meet different cohorts at very different ages.

```{r}
#| label: prepare-loans

country_code <- "EE"
cohort_from <- as.Date("2018-01-01")
cohort_to <- as.Date("2022-12-01")
max_age <- 36L

# Excel exports can store dates as date cells, timestamps or serial numbers.
as_bondora_date <- function(x) {
  if (inherits(x, "Date")) return(x)
  if (inherits(x, "POSIXt")) return(as.Date(x))
  if (is.numeric(x)) return(openxlsx::convertToDate(x))
  as.Date(substr(as.character(x), 1, 10))
}

loans <- loans_raw |>
  select(
    loan_id = LoanId,
    country = Country,
    status = Status,
    loan_date = LoanDate,
    default_date = DefaultDate,
    contract_end_date = ContractEndDate,
    last_payment_date = LastPaymentOn,
    report_date = ReportAsOfEOD
  ) |>
  mutate(
    loan_date = as_bondora_date(loan_date),
    default_date = as_bondora_date(default_date),
    contract_end_date = as_bondora_date(contract_end_date),
    last_payment_date = as_bondora_date(last_payment_date),
    report_date = as_bondora_date(report_date),
    cohort = floor_date(loan_date, unit = "month"),
    default_month = floor_date(default_date, unit = "month"),
    contract_end_month = floor_date(contract_end_date, unit = "month"),
    last_payment_month = floor_date(last_payment_date, unit = "month"),
    report_month = floor_date(report_date, unit = "month")
  ) |>
  filter(
    country == country_code,
    cohort >= cohort_from,
    cohort <= cohort_to
  ) |>
  mutate(
    # Default ends the history. Otherwise we stop at repayment or the reporting month.
    exit_month = case_when(
      !is.na(default_month) ~ default_month,
      status == "Repaid" & !is.na(last_payment_month) ~ last_payment_month,
      status == "Repaid" & !is.na(contract_end_month) ~ contract_end_month,
      TRUE ~ report_month
    ),
    exit_age = interval(cohort, exit_month) %/% months(1),
    default_age = if_else(
      !is.na(default_month),
      interval(cohort, default_month) %/% months(1),
      NA_real_
    ),
    observed_age = pmin(exit_age, max_age)
  ) |>
  filter(
    observed_age >= 1,
    is.na(default_age) | default_age >= 1
  )

loans
```

At this point there is still one row per loan. We know when it started and, when
it happened, when default occurred. We do not yet have one row per observed month.

## One loan, month by month

Start with one loan that eventually defaulted.

```{r}
#| label: choose-one-loan

one_loan <- loans |>
  filter(!is.na(default_age), default_age <= max_age) |>
  arrange(abs(default_age - 6)) |>
  slice(1) |>
  select(loan_id, cohort, observed_age, default_age)

one_loan
```

`uncount()` turns that one record into one row per observed month. Age 1 is the
first month after origination. The default flag is one only in the month when the
default first appears.

```{r}
#| label: expand-one-loan

one_loan_months <- one_loan |>
  uncount(observed_age, .id = "age") |>
  mutate(
    period = cohort %m+% months(age),
    default = as.integer(!is.na(default_age) & age == default_age)
  ) |>
  select(loan_id, cohort, age, period, default)

one_loan_months
```

A loan defaulting at age 6 appears in six monthly rows, but contributes exactly
one new default: at age 6. We do not create rows after that default.

## The same transformation for every loan

Now put the same logic in a small function.

```{r}
#| label: define-expand-loan

expand_loan <- function(loan_id, cohort, observed_age, default_age) {
  tibble::tibble(
    loan_id = loan_id,
    cohort = cohort,
    age = seq_len(as.integer(observed_age))
  ) |>
    mutate(
      period = cohort %m+% months(age),
      default = as.integer(!is.na(default_age) & age == default_age)
    )
}
```

`pmap()` applies that function to every loan and `list_rbind()` joins the results.

```{r}
#| label: expand-all-loans

loan_months <- loans |>
  select(loan_id, cohort, observed_age, default_age) |>
  purrr::pmap(expand_loan) |>
  purrr::list_rbind()

loan_months
```

We have moved from one row per loan to:

`loan_id × cohort × age × period × default`.

## From monthly records to vintage hazards

Now group those rows by cohort, age and period. For each combination we count how
many loans reached the month and how many new defaults happened there.

```{r}
#| label: build-vintage-long

vintage_long <- loan_months |>
  summarise(
    exposure = n(),
    defaults = sum(default),
    .by = c(cohort, age, period)
  ) |>
  mutate(hazard = defaults / exposure) |>
  arrange(cohort, age)

vintage_long
```

For each row:

\[
\text{hazard} =
\frac{\text{new defaults during the month}}
     {\text{loans exposed at the start of the month}}.
\]

So the hazard is simply the monthly default rate among loans that reached that
month without defaulting before. It is not an accumulated default rate.

The long table is the object we will use for the rest of the analysis. Each row has
three identities at once. A hazard can be high because of the age of the loans,
because that cohort is different, because something happened in that calendar
period, or because of a mixture of those things.

## The familiar vintage triangle

The same long table can be displayed with cohort down the rows and age across the
columns. Here we take twelve cohorts around 2020 and their first twelve months.

```{r}
#| label: show-vintage-triangle
#| column: page

triangle_cohorts <- seq.Date(
  as.Date("2019-06-01"),
  as.Date("2020-05-01"),
  by = "month"
)

vintage_triangle <- vintage_long |>
  filter(
    cohort %in% triangle_cohorts,
    age <= 12
  ) |>
  select(cohort, age, hazard) |>
  mutate(
    cohort = format(cohort, "%Y-%m"),
    hazard = scales::percent(hazard, accuracy = 0.01)
  ) |>
  pivot_wider(
    names_from = age,
    names_prefix = "M",
    values_from = hazard
  ) |>
  arrange(cohort)

vintage_triangle
```

The triangle is only another view of `vintage_long`; it is not a different data
source. It also shows why the three clocks are mixed. A calendar month cuts across
the triangle diagonally: loans from different cohorts reach the same period at
different ages.

For example, look at April 2020 directly in the long table:

```{r}
#| label: inspect-one-period

period_example <- as.Date("2020-04-01")

vintage_long |>
  filter(
    period == period_example,
    cohort >= as.Date("2018-01-01"),
    cohort <= as.Date("2020-03-01")
  ) |>
  select(cohort, age, period, exposure, defaults, hazard) |>
  arrange(cohort)
```

It is the **same period**, but the 2018 loans are much older than the 2019 loans,
and the early-2020 loans are only a few months old. If risk changes around that
period, a raw vintage curve alone cannot tell us whether the difference comes from
age, cohort or the common calendar period.

That is why including pre-2020 cohorts is useful here. If the PERIOD component
later changes around 2020, it is a pattern worth inspecting. The decomposition by
itself does not prove that COVID caused the change; it only separates the calendar
pattern from AGE and COHORT as far as this method allows.

## Cumulative default by cohort

Before decomposing anything, look at a classic vintage view: cumulative default as
cohorts get older. January cohorts from 2018 to 2022 give us loans that encountered
2020 at very different ages, plus cohorts that started after it.

```{r}
#| label: plot-cumulative-vintages
#| column: page

cumulative_vintages <- vintage_long |>
  group_by(cohort) |>
  arrange(age, .by_group = TRUE) |>
  mutate(
    initial_loans = first(exposure),
    cumulative_defaults = cumsum(defaults),
    cumulative_default_rate = cumulative_defaults / initial_loans
  ) |>
  ungroup() |>
  mutate(cohort_label = format(cohort, "%Y-%m"))

p_vintages <- cumulative_vintages |>
  filter(month(cohort) == 1) |>
  ggplot(aes(age, cumulative_default_rate, colour = cohort_label)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
  labs(
    title = "Cumulative default by cohort",
    subtitle = "January cohorts from 2018 to 2022",
    x = "Age (months)",
    y = "Cumulative default rate",
    colour = "Cohort"
  )

p_vintages
```

Different lines can have different shapes for many reasons. The loans were
originated in different months, they reached each age in different calendar
periods, and they passed through events such as 2020 at different points in their
life. That is exactly the mixture we want to separate.

This raw vintage chart is also a natural candidate for the post preview. It is not
a signature curve: each cohort remains separate and nothing is normalised into one
common shape.

## Raw age profile

We can also ignore cohort and period for a moment and group only by age.

```{r}
#| label: plot-raw-age-hazard

age_hazard <- vintage_long |>
  summarise(
    exposure = sum(exposure),
    defaults = sum(defaults),
    .by = age
  ) |>
  mutate(hazard = defaults / exposure)

age_hazard

ggplot(age_hazard, aes(age, hazard)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1)) +
  labs(
    title = "Default hazard by loan age",
    subtitle = "Raw age profile before separating cohort and period",
    x = "Age (months)",
    y = "Monthly default hazard"
  )
```

This curve is useful, but it still mixes the three parts. Age 12 for a 2018 cohort
happened in a different period from age 12 for a 2020 cohort.

## Sequential APC decomposition

We now return to `vintage_long`. The triangle and cumulative curves were only for
looking at the data; the decomposition uses the complete long table.

Some cohort-age-period rows have no defaults. Their raw hazard is exactly zero,
and `logit(0)` is not finite. To avoid that, use a small **add-half correction**:
add half a default and half a non-default to every row.

\[
q = \frac{defaults + 0.5}{exposure + 1}.
\]

The `0.5` in the numerator is the half default. The denominator increases by `1`
because we added `0.5` to each of the two outcomes: default and no default. In
large rows the change is tiny; its main purpose is to keep the logit finite.

We then split the logit of that adjusted rate into a mean plus AGE, COHORT and
PERIOD effects:

\[
y \approx \mu + AGE + COHORT + PERIOD.
\]

The simple version used here does that in order. First AGE is estimated, then
COHORT from what remains, and finally PERIOD from what remains after both. Exposure
is used as the weight throughout.

```{r}
#| label: estimate-sequential-apc

pseudo_count <- 0.5

apc_cells <- vintage_long |>
  mutate(
    q = (defaults + pseudo_count) / (exposure + 2 * pseudo_count),
    y_logit = qlogis(q)
  )

mu <- weighted.mean(apc_cells$y_logit, apc_cells$exposure)

age_effects <- apc_cells |>
  mutate(residual = y_logit - mu) |>
  summarise(
    effect = weighted.mean(residual, exposure),
    weight = sum(exposure),
    .by = age
  ) |>
  mutate(effect = effect - weighted.mean(effect, weight))

cells_after_age <- apc_cells |>
  left_join(
    age_effects |>
      select(age, age_effect = effect),
    by = "age"
  )

cohort_effects <- cells_after_age |>
  mutate(residual = y_logit - mu - age_effect) |>
  summarise(
    effect = weighted.mean(residual, exposure),
    weight = sum(exposure),
    .by = cohort
  ) |>
  mutate(effect = effect - weighted.mean(effect, weight))

cells_after_cohort <- cells_after_age |>
  left_join(
    cohort_effects |>
      select(cohort, cohort_effect = effect),
    by = "cohort"
  )

period_effects <- cells_after_cohort |>
  mutate(residual = y_logit - mu - age_effect - cohort_effect) |>
  summarise(
    effect = weighted.mean(residual, exposure),
    weight = sum(exposure),
    .by = period
  ) |>
  mutate(effect = effect - weighted.mean(effect, weight))

apc_decomposition <- cells_after_cohort |>
  left_join(
    period_effects |>
      select(period, period_effect = effect),
    by = "period"
  ) |>
  mutate(
    fitted_logit = mu + age_effect + cohort_effect + period_effect,
    residual = y_logit - fitted_logit,
    fitted_hazard = plogis(fitted_logit),
    reconstructed_logit = fitted_logit + residual,
    reconstructed_q = plogis(reconstructed_logit)
  ) |>
  select(
    cohort,
    age,
    period,
    exposure,
    defaults,
    q,
    y_logit,
    age_effect,
    cohort_effect,
    period_effect,
    fitted_logit,
    residual,
    fitted_hazard,
    reconstructed_logit,
    reconstructed_q
  )

apc_decomposition
```

There **is** a residual. The three effects explain part of each row, not every
small fluctuation. Each row can therefore be written exactly as:

\[
y = \mu + AGE + COHORT + PERIOD + residual.
\]

`fitted_hazard` is the rate explained by the mean and the three APC effects. `q`
is the adjusted observed rate. Adding the residual back recovers the original
logit and therefore the original adjusted rate.

```{r}
#| label: check-apc-reconstruction

apc_decomposition |>
  summarise(
    max_logit_difference = max(abs(y_logit - reconstructed_logit)),
    max_rate_difference = max(abs(q - reconstructed_q))
  )
```

Those differences should be zero apart from numerical precision.

The result tables are one-dimensional: one AGE effect for every month of age, one
COHORT effect for every origination month, and one PERIOD effect for every calendar
month. The plots below put each effect back on a hazard scale. The dashed line is
the overall baseline.

```{r}
#| label: plot-age-effect

baseline_hazard <- plogis(mu)

ggplot(age_effects, aes(age, plogis(mu + effect))) +
  geom_hline(yintercept = baseline_hazard, linetype = "dashed") +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1)) +
  labs(
    title = "AGE effect",
    subtitle = "Which loan ages show more risk?",
    x = "Age (months)",
    y = "Default hazard"
  )
```

```{r}
#| label: plot-cohort-effect
#| column: page

ggplot(cohort_effects, aes(cohort, plogis(mu + effect))) +
  geom_hline(yintercept = baseline_hazard, linetype = "dashed") +
  geom_line() +
  geom_point() +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1)) +
  labs(
    title = "COHORT effect",
    subtitle = "Which origination cohorts show more risk after AGE is removed?",
    x = "Origination cohort",
    y = "Default hazard"
  )
```

```{r}
#| label: plot-period-effect
#| column: page

ggplot(period_effects, aes(period, plogis(mu + effect))) +
  geom_hline(yintercept = baseline_hazard, linetype = "dashed") +
  geom_vline(xintercept = as.Date("2020-03-01"), linetype = "dotted") +
  geom_line() +
  geom_point() +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1)) +
  labs(
    title = "PERIOD effect",
    subtitle = "The dotted line marks March 2020",
    x = "Calendar period",
    y = "Default hazard"
  )
```

The interpretation is direct. A peak in AGE points to ages where default is more
likely. A peak in COHORT points to origination months that look riskier. A peak in
PERIOD points to calendar months where the loans observed at that time look
riskier.

## One important caveat

This post uses a **sequential** decomposition because it is easy to see step by
step. But AGE, COHORT and PERIOD are linked by `period = cohort + age`, so there is
no unique way to separate all three without making an additional choice. Here the
choice is the order: AGE first, then COHORT, then PERIOD.

That is why these plots should be read as one transparent way of dividing the
observed vintage risk, not as the only possible decomposition. A later comparison
can use a simultaneous APC method and check whether the same risky ages, cohorts
and periods appear.

## References

Federal Reserve Board (2013), *Capital Planning at Large Bank Holding Companies*: vintage loss models, also called age-cohort-time models, for retail credit portfolios.  
Banco de la República (2024), *Decomposition of the Performance of Credit Vintages*: an age-period-cohort application to credit vintages in its Financial Stability Report.
