---
title: Decomposing vintage risk into age, cohort and period
description: Reconstruct monthly loan hazards from public Bondora loans and split vintage risk sequentially into age, cohort and period effects.
author: Joshua Kunst
date: 2026-09-10
categories: [R, credit risk, vintage analysis, APC]
lang: en
draft: true
toc: true
---

```{r}
#| label: setup
#| include: false

source("../../_R/post_setup.R")

install_missing_packages(c(
  "dplyr",
  "ggplot2",
  "lubridate",
  "openxlsx",
  "purrr",
  "scales",
  "tibble",
  "tidyr"
))
```

Vintage curves mix three clocks. **Age** is the number of months since a loan was
originated, **cohort** is its origination month, and **period** is the calendar
month in which performance is observed.

For monthly data they are linked exactly:

\[
\text{period} = \text{cohort} + \text{age}.
\]

The easiest way to understand that identity is to build a vintage table from the
bottom up. We will start with one loan, reconstruct its monthly history, repeat the
same operation for every loan, summarise the resulting rows into monthly hazards,
and only then introduce an AGE–COHORT–PERIOD decomposition.

## Public loan data

[Go & Grow's public statistics page](https://goandgrow.eu/en/public-statistics/)
links a downloadable Bondora loan dataset. The file contains one row per loan and
includes origination, default, contract-end and reporting dates, which is enough
to reconstruct the monthly history needed here.

The source is a live snapshot, so results can move when the provider refreshes the
file. `DefaultDate` is also Bondora's own default/collection event definition; it
should not be read as a regulatory default definition.

```{r}
#| label: download-public-loans

library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)

data_url <- "https://sabanners001.blob.core.windows.net/statistics/public/loan_dataset_investor.xlsx"

# Keep the large source file outside the repository and reuse it on later renders.
cache_dir <- tools::R_user_dir("jkunst-blog", which = "cache")
data_file <- file.path(cache_dir, "bondora-loan-dataset.xlsx")
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)

if (!file.exists(data_file)) {
  download.file(data_url, data_file, mode = "wb")
}
```

The public workbook is wide, but only a handful of fields are needed here. Reading
just those columns keeps the example lighter even as the source grows.

```{r}
#| label: read-loan-fields

needed_columns <- c(
  "ReportAsOfEOD",
  "LoanId",
  "LoanDate",
  "Country",
  "Status",
  "DefaultDate",
  "ContractEndDate",
  "LastPaymentOn"
)

# Read only the header first so column positions do not have to be hard-coded.
header <- openxlsx::read.xlsx(
  data_file,
  rows = 1,
  colNames = FALSE,
  skipEmptyCols = FALSE
) |>
  unlist(use.names = FALSE) |>
  as.character()

column_index <- match(needed_columns, header)
missing_columns <- needed_columns[is.na(column_index)]

if (length(missing_columns) > 0) {
  stop(
    "Bondora file layout changed. Missing columns: ",
    paste(missing_columns, collapse = ", ")
  )
}

loans_raw <- openxlsx::read.xlsx(
  data_file,
  cols = column_index,
  colNames = TRUE,
  detectDates = TRUE,
  check.names = FALSE
)
```

For a compact example I keep Estonian originations from 2022 through 2025 and at
most the first 36 months of performance. The recent cohorts are intentionally
included: they make the familiar incomplete vintage triangle visible.

```{r}
#| label: prepare-loans

country_code <- "EE"
cohort_from <- as.Date("2022-01-01")
cohort_to <- as.Date("2025-12-01")
max_age <- 36L

# Excel exports can store dates as date cells, timestamps or serial numbers.
as_bondora_date <- function(x) {
  if (inherits(x, "Date")) return(x)
  if (inherits(x, "POSIXt")) return(as.Date(x))
  if (is.numeric(x)) return(openxlsx::convertToDate(x))
  as.Date(substr(as.character(x), 1, 10))
}

loans <- loans_raw |>
  transmute(
    loan_id = LoanId,
    country = Country,
    status = Status,
    loan_date = as_bondora_date(LoanDate),
    default_date = as_bondora_date(DefaultDate),
    contract_end_date = as_bondora_date(ContractEndDate),
    last_payment_date = as_bondora_date(LastPaymentOn),
    report_date = as_bondora_date(ReportAsOfEOD)
  ) |>
  mutate(
    cohort = floor_date(loan_date, unit = "month"),
    default_month = floor_date(default_date, unit = "month"),
    contract_end_month = floor_date(contract_end_date, unit = "month"),
    last_payment_month = floor_date(last_payment_date, unit = "month"),
    report_month = floor_date(report_date, unit = "month")
  ) |>
  filter(
    country == country_code,
    cohort >= cohort_from,
    cohort <= cohort_to
  ) |>
  mutate(
    # Default is the event. Otherwise repayment or the reporting date censors observation.
    exit_month = case_when(
      !is.na(default_month) ~ default_month,
      status == "Repaid" & !is.na(last_payment_month) ~ last_payment_month,
      status == "Repaid" & !is.na(contract_end_month) ~ contract_end_month,
      TRUE ~ report_month
    ),
    exit_age = interval(cohort, exit_month) %/% months(1),
    default_age = if_else(
      !is.na(default_month),
      interval(cohort, default_month) %/% months(1),
      NA_real_
    ),
    observed_age = pmin(exit_age, max_age)
  ) |>
  filter(
    observed_age >= 1,
    is.na(default_age) | default_age >= 1
  )

loans |>
  summarise(
    loans = n(),
    defaults = sum(!is.na(default_age) & default_age <= max_age),
    first_cohort = min(cohort),
    last_cohort = max(cohort)
  )
```

At this point there is still one row per loan. We know when the loan started and,
if applicable, when default occurred, but there is not yet one row for every month
that the loan was exposed.

## One loan, month by month

Start with one observed default. Picking a single loan makes the expansion easy to
see before doing exactly the same thing thousands of times.

```{r}
#| label: choose-one-loan

one_loan <- loans |>
  filter(!is.na(default_age), default_age <= max_age) |>
  arrange(abs(default_age - 6)) |>
  slice(1) |>
  select(loan_id, cohort, observed_age, default_age)

one_loan
```

`uncount()` turns that one record into one row per month of observed life. Age 1 is
the first month after origination, so its calendar period is the cohort month plus
one month. The default indicator is one only in the month in which default first
appears.

```{r}
#| label: expand-one-loan

one_loan_months <- one_loan |>
  uncount(observed_age, .id = "age") |>
  mutate(
    period = cohort %m+% months(age),
    default = as.integer(!is.na(default_age) & age == default_age)
  ) |>
  select(loan_id, cohort, age, period, default)

one_loan_months
```

The transformation is the important part. A loan defaulting at age 6 appears in
six monthly risk sets, but contributes exactly **one** new default: at age 6.
Months after that event do not exist in the reconstructed history.

## The same transformation for every loan

Now put that logic in a small function. It takes the four pieces of information
that describe an observed loan and returns its monthly rows.

```{r}
#| label: define-expand-loan

expand_loan <- function(loan_id, cohort, observed_age, default_age) {
  tibble::tibble(
    loan_id = loan_id,
    cohort = cohort,
    age = seq_len(as.integer(observed_age))
  ) |>
    mutate(
      period = cohort %m+% months(age),
      default = as.integer(!is.na(default_age) & age == default_age)
    )
}
```

`pmap()` applies the same function row by row. Each loan returns a small monthly
table; `list_rbind()` joins those tables into the longitudinal portfolio history.

```{r}
#| label: expand-all-loans

loan_months <- loans |>
  select(loan_id, cohort, observed_age, default_age) |>
  purrr::pmap(expand_loan) |>
  purrr::list_rbind()

loan_months |>
  arrange(loan_id, age) |>
  slice_head(n = 12)
```

We have moved from **one row per loan** to the structure normally available in a
monthly credit portfolio:

`loan_id × cohort × age × period × default`.

## From monthly records to vintage hazards

The loan-level history is now aggregated by the three clocks. For each cell we
count the loans exposed to default during the interval and how many first defaults
occurred in that interval.

```{r}
#| label: build-vintage-long

vintage_long <- loan_months |>
  summarise(
    exposure = n(),
    defaults = sum(default),
    .by = c(cohort, age, period)
  ) |>
  mutate(hazard = defaults / exposure) |>
  arrange(cohort, age)

vintage_long |>
  slice_head(n = 15)
```

For a cell identified by cohort \(c\), age \(a\) and period \(p\):

\[
\text{hazard}_{c,a,p} =
\frac{\text{new defaults}_{c,a,p}}
     {\text{loans exposed at the start of the interval}_{c,a,p}}.
\]

This is a **conditional monthly default rate**: the probability of default during
that month given that the loan survived to the beginning of the interval. It is
not a cumulative default rate.

The long table is the object we will use for the rest of the analysis. Each row has
three identities at once. For example, one observed hazard can belong to age 4,
the January 2023 cohort, and May 2023 calendar time. APC will later ask how much of
that risk looks like maturity, origination quality or calendar environment.

## The familiar vintage triangle

A vintage table is often displayed with cohort down the rows and age across the
columns. That wide shape is useful for inspection, so take only the latest twelve
cohorts and the first twelve months.

```{r}
#| label: show-vintage-triangle

triangle_cohorts <- vintage_long |>
  distinct(cohort) |>
  arrange(desc(cohort)) |>
  slice_head(n = 12) |>
  pull(cohort)

vintage_triangle <- vintage_long |>
  filter(
    cohort %in% triangle_cohorts,
    age <= 12
  ) |>
  transmute(
    cohort = format(cohort, "%Y-%m"),
    age,
    hazard = scales::percent(hazard, accuracy = 0.01)
  ) |>
  pivot_wider(
    names_from = age,
    names_prefix = "M",
    values_from = hazard
  ) |>
  arrange(cohort)

vintage_triangle
```

The empty cells for the newest cohorts are not missing performance that should be
filled mechanically: those loans simply have not reached those ages yet.

Most importantly, **the triangle is only a view**. We do not feed this wide table
to the model. The source of truth remains `vintage_long`, where cohort, age,
period, exposure, defaults and hazard stay explicit columns.

## Raw age profile

Before separating the clocks, collapse the long table only by age. This answers a
simple descriptive question: among all loans that reached each age, how often did
a new default occur during that month?

```{r}
#| label: plot-raw-age-hazard

age_hazard <- vintage_long |>
  summarise(
    exposure = sum(exposure),
    defaults = sum(defaults),
    .by = age
  ) |>
  mutate(hazard = defaults / exposure)

ggplot(age_hazard, aes(age, hazard)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1)) +
  labs(
    title = "Default hazard by loan age",
    subtitle = "Raw maturity profile before separating cohort and calendar effects",
    x = "Age (months)",
    y = "Monthly default hazard"
  )
```

This curve is useful, but it is still confounded. Age 6 for an old cohort happened
in a different calendar month from age 6 for a new cohort. A bad origination
vintage can also lift the apparent risk at the ages where it contributes most
exposure.

### A related path: signature curves

The same vintage data can support a different question. A **signature curve**
typically follows cumulative default emergence by age, combines comparable
cohorts or segments, and normalises that cumulative curve at a chosen anchor such
as month 12. Its job is to describe the *shape through which lifetime risk
emerges* and then scale a point-in-time PD across future months.

That is not what we do here. The raw age profile above is not a signature curve,
and the AGE component below will not be one either. APC asks a different question:
**which of the three clocks is associated with the risk observed in each vintage
cell?**

## Sequential APC decomposition

We now return to `vintage_long`. The wide triangle was only for looking at the
data; the decomposition is estimated from the complete long table.

First smooth the cell hazard slightly before taking logits so zero-default cells
remain finite:

\[
q = \frac{defaults + 0.5}{exposure + 1},
\qquad
 y = \operatorname{logit}(q).
\]

Then allocate the logit variation in order:

\[
y_{a,c,p} \approx \mu + AGE_a + COHORT_c + PERIOD_p.
\]

`AGE` is estimated first around the global weighted mean. `COHORT` is estimated
from the residual after removing AGE, and `PERIOD` from the residual left after
removing both. Exposure is used as the weight throughout.

```{r}
#| label: estimate-sequential-apc

alpha <- 0.5
beta <- 1

apc_cells <- vintage_long |>
  mutate(
    q = (defaults + alpha) / (exposure + beta),
    y_logit = qlogis(q)
  )

mu <- weighted.mean(apc_cells$y_logit, apc_cells$exposure)

age_effects <- apc_cells |>
  mutate(residual = y_logit - mu) |>
  summarise(
    effect = weighted.mean(residual, exposure),
    weight = sum(exposure),
    .by = age
  ) |>
  mutate(effect = effect - weighted.mean(effect, weight))

cells_after_age <- apc_cells |>
  left_join(
    age_effects |>
      select(age, age_effect = effect),
    by = "age"
  )

cohort_effects <- cells_after_age |>
  mutate(residual = y_logit - mu - age_effect) |>
  summarise(
    effect = weighted.mean(residual, exposure),
    weight = sum(exposure),
    .by = cohort
  ) |>
  mutate(effect = effect - weighted.mean(effect, weight))

cells_after_cohort <- cells_after_age |>
  left_join(
    cohort_effects |>
      select(cohort, cohort_effect = effect),
    by = "cohort"
  )

period_effects <- cells_after_cohort |>
  mutate(residual = y_logit - mu - age_effect - cohort_effect) |>
  summarise(
    effect = weighted.mean(residual, exposure),
    weight = sum(exposure),
    .by = period
  ) |>
  mutate(effect = effect - weighted.mean(effect, weight))

apc_decomposition <- cells_after_cohort |>
  left_join(
    period_effects |>
      select(period, period_effect = effect),
    by = "period"
  ) |>
  mutate(
    fitted_logit = mu + age_effect + cohort_effect + period_effect,
    fitted_hazard = plogis(fitted_logit)
  )
```

The three result tables are now one-dimensional: one AGE effect per month of life,
one COHORT effect per origination vintage, and one PERIOD effect per calendar
month. Each effect is shown below on a probability scale as `plogis(mu + effect)`.
The dashed line is the baseline `plogis(mu)`.

```{r}
#| label: plot-age-effect

baseline_hazard <- plogis(mu)

ggplot(age_effects, aes(age, plogis(mu + effect))) +
  geom_hline(yintercept = baseline_hazard, linetype = "dashed") +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1)) +
  labs(
    title = "AGE effect",
    subtitle = "Which months of loan life carry more risk?",
    x = "Age (months)",
    y = "Hazard with other effects at zero"
  )
```

```{r}
#| label: plot-cohort-effect
#| column: page

ggplot(cohort_effects, aes(cohort, plogis(mu + effect))) +
  geom_hline(yintercept = baseline_hazard, linetype = "dashed") +
  geom_line() +
  geom_point() +
  scale_x_date(date_breaks = "6 months", date_labels = "%Y-%m") +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1)) +
  labs(
    title = "COHORT effect",
    subtitle = "Which origination vintages look riskier after removing AGE?",
    x = "Origination cohort",
    y = "Hazard with other effects at zero"
  )
```

```{r}
#| label: plot-period-effect
#| column: page

ggplot(period_effects, aes(period, plogis(mu + effect))) +
  geom_hline(yintercept = baseline_hazard, linetype = "dashed") +
  geom_line() +
  geom_point() +
  scale_x_date(date_breaks = "6 months", date_labels = "%Y-%m") +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1)) +
  labs(
    title = "PERIOD effect",
    subtitle = "Which calendar months look riskier after removing AGE and COHORT?",
    x = "Calendar period",
    y = "Hazard with other effects at zero"
  )
```

These plots are easier to read than the original triangle because each one holds
one clock fixed as its own axis. A peak in AGE points to a risky maturity zone, a
peak in COHORT to a comparatively weak origination vintage, and a peak in PERIOD
to a calendar interval that was adverse across the loans observed then — always
under the sequential allocation used here.

## One important caveat

This is intentionally the **sequential** decomposition. Because monthly APC has
the identity `period = cohort + age`, the three linear trends are not separately
identified without an additional convention. The extraction order therefore
matters here: AGE gets first claim on the variation, COHORT gets the residual, and
PERIOD gets what remains.

That does not make the plots useless. It makes their interpretation precise: they
show how this particular transparent allocation divides the observed vintage
risk. A natural next benchmark is to fit the same `vintage_long` table with a
simultaneous identified APC specification and check which risky ages, cohorts and
periods are stable across both approaches.
