---
title: Decomposing vintage risk into age, cohort and period
description: Reconstruct monthly loan hazards from public Bondora vintages and split them sequentially into age, cohort and period effects.
author: Joshua Kunst
date: 2026-09-10
categories: [R, credit risk, vintage analysis, APC]
lang: en
draft: true
toc: true
---

```{r}
#| label: setup
#| include: false

source("../../_R/post_setup.R")

install_missing_packages(c(
  "dplyr",
  "ggplot2",
  "lubridate",
  "openxlsx",
  "scales",
  "tidyr"
))
```

Vintage curves mix three clocks. **Age** is the number of months since a loan was
originated, **cohort** is its origination month, and **period** is the calendar
month in which performance is observed.

For monthly data they are linked exactly:

\[
\text{period} = \text{cohort} + \text{age}.
\]

This small example starts from public loan-level data, rebuilds the monthly risk
set, calculates the default hazard, and then reproduces a simple sequential
**AGE → COHORT → PERIOD** decomposition.

## Public loan data

[Go & Grow's public statistics page](https://goandgrow.eu/en/public-statistics/)
links a downloadable Bondora loan dataset. The file contains one row per loan and
includes origination, default, contract-end and reporting dates, which is enough
to reconstruct monthly exposure without starting from a pre-built vintage table.

The source is a live snapshot, so results can move when the provider refreshes the
file. `DefaultDate` is also Bondora's own default/collection event definition; it
should not be read as a regulatory default definition.

```{r}
#| label: download-public-loans

library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)

data_url <- "https://sabanners001.blob.core.windows.net/statistics/public/loan_dataset_investor.xlsx"

# Keep the large source file outside the repository and reuse it on later renders.
cache_dir <- tools::R_user_dir("jkunst-blog", which = "cache")
data_file <- file.path(cache_dir, "bondora-loan-dataset.xlsx")
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)

if (!file.exists(data_file)) {
  download.file(data_url, data_file, mode = "wb")
}
```

The public workbook is wide, but only a handful of fields are needed here. Reading
just those columns keeps the example lighter even as the source grows.

```{r}
#| label: read-loan-fields

needed_columns <- c(
  "ReportAsOfEOD",
  "LoanId",
  "LoanDate",
  "Country",
  "Status",
  "DefaultDate",
  "ContractEndDate",
  "LastPaymentOn"
)

# Read only the header first so column positions do not have to be hard-coded.
header <- openxlsx::read.xlsx(
  data_file,
  rows = 1,
  colNames = FALSE,
  skipEmptyCols = FALSE
) |>
  unlist(use.names = FALSE) |>
  as.character()

column_index <- match(needed_columns, header)
missing_columns <- needed_columns[is.na(column_index)]

if (length(missing_columns) > 0) {
  stop(
    "Bondora file layout changed. Missing columns: ",
    paste(missing_columns, collapse = ", ")
  )
}

loans_raw <- openxlsx::read.xlsx(
  data_file,
  cols = column_index,
  colNames = TRUE,
  detectDates = TRUE,
  check.names = FALSE
)
```

For a compact example I keep Estonian originations from 2018 through 2022 and the
first 36 months of performance. These parameters are deliberately visible so the
window can be changed without touching the rest of the workflow.

```{r}
#| label: prepare-loan-vintages

country_code <- "EE"
cohort_from <- as.Date("2018-01-01")
cohort_to <- as.Date("2022-12-01")
max_age <- 36L

# Excel exports can store dates either as real date cells, timestamps or serial numbers.
as_bondora_date <- function(x) {
  if (inherits(x, "Date")) return(x)
  if (inherits(x, "POSIXt")) return(as.Date(x))
  if (is.numeric(x)) return(openxlsx::convertToDate(x))
  as.Date(substr(as.character(x), 1, 10))
}

loans <- loans_raw |>
  transmute(
    loan_id = LoanId,
    country = Country,
    status = Status,
    loan_date = as_bondora_date(LoanDate),
    default_date = as_bondora_date(DefaultDate),
    contract_end_date = as_bondora_date(ContractEndDate),
    last_payment_date = as_bondora_date(LastPaymentOn),
    report_date = as_bondora_date(ReportAsOfEOD)
  ) |>
  mutate(
    cohort = floor_date(loan_date, unit = "month"),
    default_month = floor_date(default_date, unit = "month"),
    contract_end_month = floor_date(contract_end_date, unit = "month"),
    last_payment_month = floor_date(last_payment_date, unit = "month"),
    report_month = floor_date(report_date, unit = "month")
  ) |>
  filter(
    country == country_code,
    cohort >= cohort_from,
    cohort <= cohort_to
  ) |>
  mutate(
    # A default is the event. Repaid loans are censored when they end; open loans at the snapshot date.
    exit_month = case_when(
      !is.na(default_month) ~ default_month,
      status == "Repaid" & !is.na(contract_end_month) ~ contract_end_month,
      status == "Repaid" & !is.na(last_payment_month) ~ last_payment_month,
      TRUE ~ report_month
    ),
    exit_age = interval(cohort, exit_month) %/% months(1),
    default_age = if_else(
      !is.na(default_month),
      interval(cohort, default_month) %/% months(1),
      NA_real_
    ),
    observed_age = pmin(exit_age, max_age)
  ) |>
  filter(
    observed_age >= 1,
    is.na(default_age) | default_age >= 1
  )

loans |>
  summarise(
    loans = n(),
    defaults = sum(!is.na(default_age) & default_age <= max_age),
    first_cohort = min(cohort),
    last_cohort = max(cohort)
  )
```

## From loans to a vintage table

Each loan is expanded only over the months in which it is still exposed. A loan
that defaults at age 7 contributes to the risk set from age 1 through age 7 and
contributes one default in that final month. A repaid loan contributes exposure
until its observed exit month but never enters the default numerator.

```{r}
#| label: build-monthly-risk-set

loan_months <- loans |>
  select(loan_id, cohort, default_age, observed_age) |>
  uncount(observed_age, .id = "age") |>
  mutate(
    period = cohort %m+% months(age),
    default = as.integer(!is.na(default_age) & age == default_age)
  )

apc_cells <- loan_months |>
  summarise(
    exposure = n(),
    defaults = sum(default),
    .by = c(cohort, age, period)
  ) |>
  mutate(hazard = defaults / exposure)

apc_cells |>
  arrange(period, cohort, age) |>
  slice_head(n = 12)
```

The raw age curve is now just the number of defaults occurring at each age divided
by the number of loans still exposed at the start of that monthly interval.

```{r}
#| label: plot-raw-age-hazard

age_hazard <- apc_cells |>
  summarise(
    exposure = sum(exposure),
    defaults = sum(defaults),
    .by = age
  ) |>
  mutate(hazard = defaults / exposure)

ggplot(age_hazard, aes(age, hazard)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1)) +
  labs(
    title = "Default hazard by loan age",
    subtitle = "Raw vintage view before separating cohort and calendar effects",
    x = "Age (months)",
    y = "Monthly default hazard"
  )
```

## Sequential APC decomposition

The earlier APC workflow can be reproduced directly on these cells. I smooth the
cell hazard slightly before taking logits so zero-default cells remain finite:

\[
q = \frac{defaults + 0.5}{exposure + 1},
\qquad
 y = \operatorname{logit}(q).
\]

Then the decomposition allocates the logit variation in order:

\[
y_{a,c,p} \approx \mu + AGE_a + COHORT_c + PERIOD_p.
\]

`AGE` is estimated first around the global weighted mean. `COHORT` is estimated
from the residual after removing AGE, and `PERIOD` from the residual left after
removing both.

```{r}
#| label: estimate-sequential-apc

alpha <- 0.5
beta <- 1

apc_cells <- apc_cells |>
  mutate(
    q = (defaults + alpha) / (exposure + beta),
    y_logit = qlogis(q)
  )

mu <- weighted.mean(apc_cells$y_logit, apc_cells$exposure)

age_effects <- apc_cells |>
  mutate(residual = y_logit - mu) |>
  summarise(
    effect = weighted.mean(residual, exposure),
    weight = sum(exposure),
    .by = age
  ) |>
  mutate(effect = effect - weighted.mean(effect, weight))

cells_after_age <- apc_cells |>
  left_join(
    age_effects |>
      select(age, age_effect = effect),
    by = "age"
  )

cohort_effects <- cells_after_age |>
  mutate(residual = y_logit - mu - age_effect) |>
  summarise(
    effect = weighted.mean(residual, exposure),
    weight = sum(exposure),
    .by = cohort
  ) |>
  mutate(effect = effect - weighted.mean(effect, weight))

cells_after_cohort <- cells_after_age |>
  left_join(
    cohort_effects |>
      select(cohort, cohort_effect = effect),
    by = "cohort"
  )

period_effects <- cells_after_cohort |>
  mutate(residual = y_logit - mu - age_effect - cohort_effect) |>
  summarise(
    effect = weighted.mean(residual, exposure),
    weight = sum(exposure),
    .by = period
  ) |>
  mutate(effect = effect - weighted.mean(effect, weight))

apc_decomposition <- cells_after_cohort |>
  left_join(
    period_effects |>
      select(period, period_effect = effect),
    by = "period"
  ) |>
  mutate(
    fitted_logit = mu + age_effect + cohort_effect + period_effect,
    fitted_hazard = plogis(fitted_logit)
  )
```

The component plots put each effect back on a probability scale as
`plogis(mu + effect)`. The dashed line is the baseline `plogis(mu)`, so values
above it identify ages, vintages or calendar periods allocated more risk by the
sequential decomposition.

```{r}
#| label: plot-age-effect

baseline_hazard <- plogis(mu)

ggplot(age_effects, aes(age, plogis(mu + effect))) +
  geom_hline(yintercept = baseline_hazard, linetype = "dashed") +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1)) +
  labs(
    title = "AGE effect",
    subtitle = "Which months of loan life carry more risk?",
    x = "Age (months)",
    y = "Hazard with other effects at zero"
  )
```

```{r}
#| label: plot-cohort-effect
#| column: page

ggplot(cohort_effects, aes(cohort, plogis(mu + effect))) +
  geom_hline(yintercept = baseline_hazard, linetype = "dashed") +
  geom_line() +
  geom_point() +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1)) +
  labs(
    title = "COHORT effect",
    subtitle = "Which origination vintages look riskier after removing AGE?",
    x = "Origination cohort",
    y = "Hazard with other effects at zero"
  )
```

```{r}
#| label: plot-period-effect
#| column: page

ggplot(period_effects, aes(period, plogis(mu + effect))) +
  geom_hline(yintercept = baseline_hazard, linetype = "dashed") +
  geom_line() +
  geom_point() +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1)) +
  labs(
    title = "PERIOD effect",
    subtitle = "Which calendar months look riskier after removing AGE and COHORT?",
    x = "Calendar period",
    y = "Hazard with other effects at zero"
  )
```

## One important caveat

This is intentionally the **sequential** decomposition. Because monthly APC has
the identity `period = cohort + age`, the three linear trends are not separately
identified without an additional convention. The extraction order therefore
matters here: AGE gets first claim on the variation, COHORT gets the residual, and
PERIOD gets what remains.

That does not make the plots useless. It makes their interpretation precise: they
show how this particular transparent allocation divides the observed vintage
risk. A natural next benchmark is to fit the same `apc_cells` with a simultaneous
identified APC specification and check which risky ages, cohorts and periods are
stable across both approaches.
