---
title: Decomposing vintage risk into age, cohort and period
description: Reconstruct monthly loan hazards from public Bondora loans and split vintage risk into age, cohort and period effects.
author: Joshua Kunst
date: 2026-09-10
categories: [R, credit risk, vintage analysis, APC]
lang: en
draft: true
toc: true
---

```{r}
#| label: setup
#| include: false

source("blog/_R/post_setup.R")

install_missing_packages(c(
  "dplyr",
  "ggplot2",
  "knitr",
  "lubridate",
  "openxlsx",
  "scales",
  "tibble",
  "tidyr"
))
```

Credit vintages are a common way to follow how default changes after loans are
originated. The [Federal Reserve](https://www.federalreserve.gov/bankinforeg/bcreg20130819a1.pdf)
describes vintage loss models, also called **age-cohort-time models**, for retail
credit portfolios. The [Banco de la República of Colombia](https://www.banrep.gov.co/en/publications-research/financial-stability-report/box-1-second-half-2024)
uses an **age-period-cohort decomposition** directly on credit vintages to help
understand where changes in credit risk come from.

The idea is simple. Every observation in a vintage has three dates attached to it:

- **Age**: how many months old the loan is.
- **Cohort**: the month when the loan was originated.
- **Period**: the calendar month when we observe it.

For monthly data they are linked exactly:

$$
\text{period} = \text{cohort} + \text{age}.
$$

We will build the data from the bottom up: one loan, its monthly history, all loans,
a classic cumulative vintage view, the monthly hazard behind it, and finally the
APC decomposition.

## Public loan data

[Go & Grow's public statistics page](https://goandgrow.eu/en/public-statistics/)
links a downloadable Bondora loan dataset. The workbook contains one row per loan
and a second sheet with a data dictionary.

```{r}
#| label: download-public-loans

library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)

data_url <- "https://sabanners001.blob.core.windows.net/statistics/public/loan_dataset_investor.xlsx"

# Keep the large source file outside the repository and reuse it on later renders.
cache_dir <- tools::R_user_dir("jkunst-blog", which = "cache")
data_file <- file.path(cache_dir, "bondora-loan-dataset.xlsx")
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)

if (!file.exists(data_file)) {
  download.file(data_url, data_file, mode = "wb")
}
```

We only need six fields from the 31 columns in the loan sheet. The names below
come directly from the current public workbook.

```{r}
#| label: read-loan-fields

needed_columns <- c(
  "loan_id",
  "country",
  "loan_issued_at",
  "loan_status",
  "is_default",
  "months_on_book"
)

# Read row 1 first so we can select only the columns needed from the large workbook.
header <- openxlsx::read.xlsx(
  data_file,
  sheet = "Loan Dataset",
  rows = 1,
  colNames = FALSE,
  skipEmptyCols = FALSE
) |>
  unlist(use.names = FALSE) |>
  as.character()

column_index <- match(needed_columns, header)

if (anyNA(column_index)) stop("Missing Bondora fields: ", paste(needed_columns[is.na(column_index)], collapse = ", "))

loans_raw <- openxlsx::read.xlsx(
  data_file,
  sheet = "Loan Dataset",
  cols = column_index,
  detectDates = TRUE,
  check.names = FALSE
) |>
  tibble::as_tibble()

loans_raw
```

Printing the tibble is intentional: it shows the number of rows and columns, the
fields we loaded and a few real observations before we transform anything.

The data contains loans from several countries. Before choosing one for the
example, look at their sizes.

```{r}
#| label: count-countries

country_counts <- loans_raw |>
  count(country, sort = TRUE)

country_counts
```

For the decomposition we use Estonia. It is not the largest country in the current
file, but it has substantial and continuous monthly originations throughout the
2018--2022 window. That makes it a cleaner example than markets with long gaps in
origination, while keeping country differences out of the AGE, COHORT and PERIOD
effects.

The workbook also explains the fields we use. In particular, `months_on_book` is
the number of months a loan has been in active status.

```{r}
#| label: inspect-data-dictionary

loan_dictionary <- openxlsx::read.xlsx(
  data_file,
  sheet = "Dataset Dictionary"
) |>
  tibble::as_tibble() |>
  filter(Column %in% needed_columns) |>
  select(Column, Description)

knitr::kable(
  loan_dictionary,
  col.names = c("Field", "Description"),
  align = c("l", "l")
)
```

For this example we keep Estonian cohorts from 2018 through 2022 and follow at
most the first 36 months. Starting before 2020 is useful because the same calendar
period can meet different cohorts at very different ages.

```{r}
#| label: prepare-loans

country_name <- "Estonia"
cohort_from <- as.Date("2018-01-01")
cohort_to <- as.Date("2022-12-01")
max_age <- 36L

loans <- loans_raw |>
  mutate(
    # Excel stores this field as a serial date in the current workbook.
    loan_date = as.Date(loan_issued_at, origin = "1899-12-30"),
    cohort = floor_date(loan_date, unit = "month"),
    months_on_book = as.integer(months_on_book),
    is_default = as.logical(is_default),
    # Use Bondora's MOB convention directly: a defaulted loan defaults at this age.
    default_age = if_else(is_default, months_on_book, NA_integer_),
    observed_age = pmin(months_on_book, max_age)
  ) |>
  filter(
    country == country_name,
    cohort >= cohort_from,
    cohort <= cohort_to,
    observed_age >= 1
  ) |>
  select(
    loan_id,
    country,
    loan_status,
    cohort,
    months_on_book,
    is_default,
    default_age,
    observed_age
  )

loans
```

At this point there is still one row per loan. We know its cohort, how many months
we observe it, and whether default happened at the end of that observed path.

## One loan, month by month

Start with one loan that defaulted around age 6.

```{r}
#| label: choose-one-loan

one_loan <- loans |>
  filter(is_default, default_age <= max_age) |>
  arrange(abs(default_age - 6)) |>
  slice(1) |>
  select(loan_id, cohort, observed_age, default_age)

one_loan
```

`uncount()` turns that one record into one row per observed month. Age 1 is the
first month after origination. The default flag is one only in the month when the
default first appears.

```{r}
#| label: expand-one-loan

one_loan_months <- one_loan |>
  uncount(observed_age, .id = "age") |>
  mutate(
    period = cohort %m+% months(age),
    default = as.integer(!is.na(default_age) & age == default_age)
  ) |>
  select(loan_id, cohort, age, period, default)

one_loan_months
```

A loan defaulting at age 6 appears in six monthly rows but contributes exactly one
new default: at age 6. We do not create rows after that default.

The exact same transformation can be applied directly to the full portfolio.

```{r}
#| label: expand-all-loans

loan_months <- loans |>
  select(loan_id, cohort, observed_age, default_age) |>
  uncount(observed_age, .id = "age") |>
  mutate(
    period = cohort %m+% months(age),
    default = as.integer(!is.na(default_age) & age == default_age)
  ) |>
  select(loan_id, cohort, age, period, default)

loan_months
```

We have moved from one row per loan to:

`loan_id × cohort × age × period × default`.

## From monthly records to vintage hazards

Now group those rows by cohort, age and period. For each combination we count how
many loans reached the month and how many new defaults happened there.

```{r}
#| label: build-vintage-long

vintage_long <- loan_months |>
  summarise(
    surviving_loans = n(),
    defaults = sum(default),
    .by = c(cohort, age, period)
  ) |>
  mutate(hazard = defaults / surviving_loans) |>
  arrange(cohort, age)

vintage_long
```

`surviving_loans` is the number of loans still present in the risk set at that age.
It is a loan count, not a monetary exposure or EAD.

For each row:

$$
\text{monthly hazard} =
\frac{\text{new defaults during the month}}
     {\text{surviving loans in that month}}.
$$

The hazard answers a local question: among loans that reached this age without
defaulting before, how many default now?

## The classic vintage view

The more familiar credit-risk vintage is cumulative. For each cohort and age we
accumulate the defaults and divide by the number of loans at the start of that
cohort.

$$
\text{cumulative default rate}_{c,a} =
\frac{\text{defaults accumulated through age } a}
     {\text{initial loans in cohort } c}.
$$

```{r}
#| label: build-cumulative-vintages

vintage_cumulative <- vintage_long |>
  arrange(cohort, age) |>
  mutate(
    initial_loans = first(surviving_loans),
    cumulative_defaults = cumsum(defaults),
    cumulative_default_rate = cumulative_defaults / initial_loans,
    .by = cohort
  )

vintage_cumulative
```

Here is the classic triangle for twelve cohorts around 2020 and their first twelve
months.

```{r}
#| label: show-vintage-triangle

triangle_cohorts <- seq.Date(
  as.Date("2019-06-01"),
  as.Date("2020-05-01"),
  by = "month"
)

vintage_triangle <- vintage_cumulative |>
  filter(
    cohort %in% triangle_cohorts,
    age <= 12
  ) |>
  select(cohort, age, cumulative_default_rate) |>
  mutate(
    cohort = format(cohort, "%Y-%m"),
    cumulative_default_rate = scales::percent(cumulative_default_rate, accuracy = 0.1)
  ) |>
  pivot_wider(
    names_from = age,
    names_prefix = "M",
    values_from = cumulative_default_rate
  ) |>
  arrange(cohort)

vintage_triangle
```

The cumulative view answers *how much of the original cohort has defaulted so far*.
The monthly hazard answers a different question: *where did new defaults occur?*
That distinction matters because the APC decomposition below works with the monthly
hazard, not with the cumulative rate.

We can display the same twelve cohorts as a monthly-hazard triangle.

```{r}
#| label: show-hazard-triangle

hazard_triangle <- vintage_long |>
  filter(
    cohort %in% triangle_cohorts,
    age <= 12
  ) |>
  select(cohort, age, hazard) |>
  mutate(
    cohort = format(cohort, "%Y-%m"),
    hazard = scales::percent(hazard, accuracy = 0.01)
  ) |>
  pivot_wider(
    names_from = age,
    names_prefix = "M",
    values_from = hazard
  ) |>
  arrange(cohort)

hazard_triangle
```

A calendar period cuts diagonally across either triangle because different cohorts
reach the same period at different ages. For example, look at April 2020 directly
in the long table:

```{r}
#| label: inspect-one-period

period_example <- as.Date("2020-04-01")

vintage_long |>
  filter(
    period == period_example,
    cohort >= as.Date("2018-01-01"),
    cohort <= as.Date("2020-03-01")
  ) |>
  select(cohort, age, period, surviving_loans, defaults, hazard) |>
  arrange(cohort)
```

It is the **same period**, but the 2018 loans are much older than the 2019 loans,
and the early-2020 loans are only a few months old. A raw vintage view cannot tell
us whether a difference comes from age, cohort or the common calendar period.

That is why including pre-2020 cohorts is useful here. If the PERIOD component
later changes around 2020, it is a pattern worth inspecting. The decomposition by
itself does not prove that COVID caused the change.

## Cumulative default by cohort

First, pool monthly cohorts into origination years for a clean overview. This is a
descriptive chart only; the APC decomposition still uses monthly cohorts.

```{r}
#| label: plot-cumulative-vintages
#| column: page

annual_vintages <- loan_months |>
  mutate(cohort_year = year(cohort)) |>
  summarise(
    surviving_loans = n(),
    defaults = sum(default),
    .by = c(cohort_year, age)
  ) |>
  arrange(cohort_year, age) |>
  mutate(
    initial_loans = first(surviving_loans),
    cumulative_defaults = cumsum(defaults),
    cumulative_default_rate = cumulative_defaults / initial_loans,
    .by = cohort_year
  )

p_vintages <- annual_vintages |>
  mutate(cohort_year = factor(cohort_year)) |>
  ggplot(aes(age, cumulative_default_rate, colour = cohort_year, group = cohort_year)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
  labs(
    title = "Cumulative default by vintage",
    subtitle = "Estonian loans originated from 2018 to 2022",
    x = "Age (months)",
    y = "Cumulative default rate",
    colour = "Cohort year"
  )

p_vintages
```

The yearly view can hide differences among monthly cohorts. Faceting by origination
year lets us inspect all twelve monthly vintages inside each year.

```{r}
#| label: plot-monthly-vintages-by-year
#| column: page

monthly_vintages <- vintage_cumulative |>
  mutate(
    cohort_year = year(cohort),
    cohort_month = month(cohort, label = TRUE, abbr = TRUE)
  )

monthly_vintages |>
  ggplot(aes(age, cumulative_default_rate, colour = cohort_month, group = cohort)) +
  geom_line() +
  facet_wrap(vars(cohort_year)) +
  scale_colour_viridis_d(begin = 0.15, end = 0.85) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
  labs(
    title = "Monthly vintages within each year",
    subtitle = "Each line is an origination month",
    x = "Age (months)",
    y = "Cumulative default rate",
    colour = "Month"
  )
```

Different lines can have different shapes because loans were originated in
different months, reached each age in different calendar periods and passed through
common events at different points in their life. That mixture is exactly what we
want to separate.

## Raw age profile

Before decomposing anything, ignore cohort and period and group only by age.

```{r}
#| label: plot-raw-age-hazard

age_hazard <- vintage_long |>
  summarise(
    surviving_loans = sum(surviving_loans),
    defaults = sum(defaults),
    .by = age
  ) |>
  mutate(hazard = defaults / surviving_loans)

age_hazard

ggplot(age_hazard, aes(age, hazard)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1)) +
  labs(
    title = "Default hazard by loan age",
    subtitle = "Raw age profile before separating cohort and period",
    x = "Age (months)",
    y = "Monthly default hazard"
  )
```

This curve is useful, but it still mixes the three parts. Age 12 for a 2018 cohort
happened in a different period from age 12 for a 2020 cohort.

## Sequential APC decomposition

We now return to `vintage_long`. Some cohort-age-period rows have no defaults.
Their raw hazard is exactly zero, and `logit(0)` is not finite. We therefore use a
small **add-half correction**: add half a default and half a non-default to every
cell.

$$
q = \frac{defaults + 0.5}{surviving\ loans + 1}.
$$

The numerator increases by `0.5`; the denominator increases by `1` because the
correction adds `0.5` to each of the two outcomes. In large cells the change is
tiny; its purpose is simply to keep the logit finite.

A probability is bounded between zero and one, so additive effects are awkward on
that scale. We therefore transform the adjusted hazard `q` to log-odds:

$$
y = \operatorname{logit}(q) = \log\left(\frac{q}{1-q}\right).
$$

After combining the effects, `plogis()` takes us back to a default probability:

$$
q = \operatorname{plogis}(y) = \frac{1}{1 + e^{-y}}.
$$

The sequential idea can be written as:

$$
y_{a,c,p} = \mu + A_a + C_{c \mid A} + P_{p \mid A,C} + \varepsilon_{a,c,p}.
$$

Read that from left to right. Start with the overall mean `mu`. AGE explains a
systematic deviation from that mean. COHORT then explains part of what remains
after AGE. PERIOD finally explains part of what remains after both AGE and COHORT.
The last term is the cell-level residual left unexplained.

### Start from the mean

Create the adjusted hazard, move it to logit space and attach the same weighted
portfolio mean to every row.

```{r}
#| label: apc-start-from-mean

apc_decomposition <- vintage_long |>
  mutate(
    q = (defaults + 0.5) / (surviving_loans + 1),
    y_logit = qlogis(q),
    mu = weighted.mean(y_logit, surviving_loans),
    residual_after_mean = y_logit - mu
  )
```

### AGE: explain the first residual

For each loan age, take the surviving-loan-weighted average of the residual left
after the mean.

```{r}
#| label: apc-add-age

apc_decomposition <- apc_decomposition |>
  mutate(
    age_effect = weighted.mean(residual_after_mean, surviving_loans),
    .by = age
  ) |>
  mutate(
    residual_after_age = y_logit - mu - age_effect
  )
```

`residual_after_age` is what the overall level and AGE still have not explained.

### COHORT: explain what AGE left behind

Now ask whether that remaining variation is systematically different across
origination cohorts.

```{r}
#| label: apc-add-cohort

apc_decomposition <- apc_decomposition |>
  mutate(
    cohort_effect = weighted.mean(residual_after_age, surviving_loans),
    .by = cohort
  ) |>
  mutate(
    residual_after_cohort = y_logit - mu - age_effect - cohort_effect
  )
```

### PERIOD: explain what AGE and COHORT left behind

Finally, group the remaining residual by calendar period. This captures common
calendar-time movements after the AGE and COHORT contributions have already been
removed.

```{r}
#| label: apc-add-period

apc_decomposition <- apc_decomposition |>
  mutate(
    period_effect = weighted.mean(residual_after_cohort, surviving_loans),
    .by = period
  ) |>
  mutate(
    residual_after_period = y_logit - mu - age_effect - cohort_effect - period_effect
  )
```

At this point the decomposition itself is finished: mean, AGE, COHORT, PERIOD and
the final residual are all explicit columns.

### Reconstruct the fitted hazard

Now combine the three effects, return from log-odds to a probability and verify
that adding the final residual reconstructs the adjusted observed rate.

```{r}
#| label: apc-reconstruct-hazard

apc_decomposition <- apc_decomposition |>
  mutate(
    fitted_logit = mu + age_effect + cohort_effect + period_effect,
    fitted_hazard = plogis(fitted_logit),
    reconstructed_q = plogis(fitted_logit + residual_after_period)
  )

glimpse(apc_decomposition)
```

The columns now tell the story of the algorithm from left to right: observed cell,
adjusted rate, logit, overall mean, residual after the mean, AGE contribution,
residual after AGE, COHORT contribution, residual after COHORT, PERIOD contribution
and final residual.

The exact row-level identity is:

$$
y = \mu + AGE + COHORT + PERIOD + residual.
$$

```{r}
#| label: check-apc-reconstruction

apc_decomposition |>
  summarise(
    max_rate_difference = max(abs(q - reconstructed_q))
  )
```

That difference should be zero apart from numerical precision.

## Reading the three effects

The decomposition is additive in log-odds. Exponentiating one effect turns it into
an odds multiplier: `1x` means no change, `1.5x` means 50% higher odds and `0.5x`
means half the odds. These are **relative effects**, not three separate default
rates.

```{r}
#| label: plot-age-effect

apc_decomposition |>
  distinct(age, age_effect) |>
  ggplot(aes(age, exp(age_effect))) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.01, suffix = "x")) +
  labs(
    title = "AGE effect",
    subtitle = "1x means average odds; above 1x means higher odds",
    x = "Age (months)",
    y = "Odds multiplier"
  )
```

```{r}
#| label: plot-cohort-effect
#| column: page

apc_decomposition |>
  distinct(cohort, cohort_effect) |>
  ggplot(aes(cohort, exp(cohort_effect))) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_line() +
  geom_point() +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.01, suffix = "x")) +
  labs(
    title = "COHORT effect",
    subtitle = "Relative odds after AGE is removed",
    x = "Origination cohort",
    y = "Odds multiplier"
  )
```

```{r}
#| label: plot-period-effect
#| column: page

apc_decomposition |>
  distinct(period, period_effect) |>
  ggplot(aes(period, exp(period_effect))) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = as.Date("2020-03-01"), linetype = "dotted") +
  geom_line() +
  geom_point() +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.01, suffix = "x")) +
  labs(
    title = "PERIOD effect",
    subtitle = "Relative odds after AGE and COHORT are removed",
    x = "Calendar period",
    y = "Odds multiplier"
  )
```

A peak in AGE means higher odds at those ages. A peak in COHORT means higher odds
for those origination months after AGE is removed. A peak in PERIOD means higher
odds in those calendar months after AGE and COHORT are removed.

## One important caveat

This post uses a **sequential** decomposition because it is easy to see step by
step. But AGE, COHORT and PERIOD are linked by `period = cohort + age`, so there is
no unique way to separate all three without making an additional choice. Here the
choice is the order: AGE first, then COHORT, then PERIOD.

That is why these plots should be read as one transparent way of dividing the
observed vintage risk, not as the only possible decomposition. A later comparison
can use a simultaneous APC method and check whether the same risky ages, cohorts
and periods appear.

### Why do the effects average to zero?

There is no extra centering step in the code. The zero weighted means appear by
construction. The portfolio mean `mu` makes `residual_after_mean` have weighted
mean zero. AGE is then the weighted group mean of that zero-mean residual, so its
weighted mean is also zero. Subtracting AGE leaves another zero-mean residual, and
the same argument repeats for COHORT and then PERIOD.

This relies on using the same rows and the same `surviving_loans` weights at every
step. We can check the property directly:

```{r}
#| label: check-zero-mean-effects

apc_decomposition |>
  summarise(
    age = weighted.mean(age_effect, surviving_loans),
    cohort = weighted.mean(cohort_effect, surviving_loans),
    period = weighted.mean(period_effect, surviving_loans)
  )
```

The three values should be zero apart from numerical precision.

### A note on monetary weights

Here `surviving_loans` is a count, so each surviving loan contributes equally to
the decomposition. In production credit-risk applications, results are often
weighted by amount, exposure or balance because a few large defaults may matter
more economically than many small ones. Bondora provides the original loan amount,
so an amount-weighted version would be a natural extension. A true balance-weighted
history would require monthly historical balances rather than the current snapshot
balance.

## References

Federal Reserve Board (2013), *Capital Planning at Large Bank Holding Companies*: vintage loss models, also called age-cohort-time models, for retail credit portfolios.  
Banco de la República (2024), *Decomposition of the Performance of Credit Vintages*: an age-period-cohort application to credit vintages in its Financial Stability Report.