---
title: Nürburgring in 3D with rayshader
description: Add real elevation to the Nürburgring and explore the circuit as an interactive 3D terrain map.
author: Joshua Kunst
date: 2026-09-10
categories: [R, maps, rayshader, OpenStreetMap]
lang: en
draft: true
toc: true
---

```{r}
#| label: setup
#| include: false

source("../../_R/post_setup.R")

install_missing_packages(c(
  "dplyr",
  "elevatr",
  "osmdata",
  "rayshader",
  "rgl",
  "sf",
  "terra"
))

# Render WebGL without requiring a visible OpenGL window during Quarto rendering.
options(rgl.useNULL = TRUE, rgl.printRglwidget = FALSE)
rgl::setupKnitr(autoprint = FALSE)
```

The [2D Nürburgring post](../2026-09-10-nurburgring-one-geom-at-a-time/)
builds the circuit from OpenStreetMap geometries with `sf` and `ggplot2`. Here we
keep that part deliberately short and focus on one new dimension: elevation.

The first goal is simple: download the track and a digital elevation model once,
cache the processed objects with the post, and place the Nürburgring directly on
top of the terrain with `rayshader`.

## Area and style

Keep the geographic frame close to the 2D version. The elevation exaggeration is
explicit because a terrain model can be geographically correct and still look
almost flat on screen.

```{r}
#| label: terrain-parameters

library(dplyr)
library(sf)

centre_lon <- 6.95
centre_lat <- 50.35
width_km <- 14
screen_ratio <- 16 / 9
height_km <- width_km / screen_ratio

forest_green <- "#47714D"
track_colour <- "#EAF1E8"
vertical_exaggeration <- 1.5
track_offset_m <- 2
interactive_max_dim <- 400

centre_utm <- sf::st_sfc(
  sf::st_point(c(centre_lon, centre_lat)),
  crs = 4326
) |>
  sf::st_transform(25832)

centre_xy <- sf::st_coordinates(centre_utm)[1, ]

# Build a rectangular frame in metres so terrain spacing and elevation use compatible units.
frame_utm <- sf::st_bbox(
  c(
    xmin = centre_xy[[1]] - width_km * 1000 / 2,
    ymin = centre_xy[[2]] - height_km * 1000 / 2,
    xmax = centre_xy[[1]] + width_km * 1000 / 2,
    ymax = centre_xy[[2]] + height_km * 1000 / 2
  ),
  crs = sf::st_crs(25832)
) |>
  sf::st_as_sfc()

frame_wgs84 <- frame_utm |>
  sf::st_transform(4326)
```

## Data

The post keeps a single processed RDS. On the first run we download raceway ways
from OpenStreetMap and elevation tiles from AWS through `elevatr`; subsequent
renders read only the local bundle.

The RDS stores the `sf` track, the elevation matrix, its projected extent and the
grid resolution. That is enough to reproduce the 3D scene without serializing a
live raster connection.

```{r}
#| label: prepare-terrain-data
#| message: false

cache_file <- "data/nurburgring-3d.rds"

if (!file.exists(cache_file)) {
  bbox_wgs84 <- sf::st_bbox(frame_wgs84)

  raceway_query <- osmdata::opq(
    bbox = bbox_wgs84,
    timeout = 180
  ) |>
    osmdata::add_osm_feature(
      key = "highway",
      value = "raceway"
    )

  track <- osmdata::osmdata_sf(raceway_query)$osm_lines |>
    filter(highway == "raceway") |>
    select(osm_id, name, geometry) |>
    sf::st_transform(25832) |>
    sf::st_intersection(frame_utm)

  # Two corner points are enough for elevatr to determine the requested tile coverage.
  # Avoid passing the frame polygon because current sf/elevatr combinations can fail there.
  elev_locations <- data.frame(
    x = c(bbox_wgs84[["xmin"]], bbox_wgs84[["xmax"]]),
    y = c(bbox_wgs84[["ymin"]], bbox_wgs84[["ymax"]])
  )

  # Let elevatr return complete AWS tiles; crop the projected raster ourselves below.
  dem <- elevatr::get_elev_raster(
    locations = elev_locations,
    z = 13,
    prj = "EPSG:4326",
    src = "aws",
    verbose = TRUE
  )

  # elevatr 0.99.x returns RasterLayer; upcoming versions return SpatRaster.
  if (!inherits(dem, "SpatRaster")) {
    dem <- methods::as(dem, "SpatRaster")
  }

  dem <- dem |>
    terra::project("EPSG:25832", method = "bilinear")

  # Crop with a plain extent after projection instead of relying on elevatr's bbox clipping.
  frame_extent <- terra::ext(terra::vect(frame_utm))
  dem <- terra::crop(dem, frame_extent)

  elevation <- rayshader::raster_to_matrix(dem)

  # render_path() can use a numeric extent in xmin, xmax, ymin, ymax order.
  terrain_extent <- as.vector(terra::ext(dem))
  names(terrain_extent) <- c("xmin", "xmax", "ymin", "ymax")

  # Elevation values are metres, so the horizontal cell size defines the natural zscale.
  resolution_m <- mean(terra::res(dem))

  dir.create("data", recursive = TRUE, showWarnings = FALSE)
  saveRDS(
    list(
      track = track,
      elevation = elevation,
      extent = terrain_extent,
      resolution_m = resolution_m
    ),
    cache_file
  )
}

terrain_data <- readRDS(cache_file)

track <- terrain_data$track
elevation <- terrain_data$elevation
terrain_extent <- terrain_data$extent
resolution_m <- terrain_data$resolution_m

c(
  elevation_min_m = min(elevation, na.rm = TRUE),
  elevation_max_m = max(elevation, na.rm = TRUE),
  grid_resolution_m = resolution_m
)
```

## Nürburgring in 3D

`plot_3d()` expects a height matrix. The cached DEM keeps its downloaded detail,
but an interactive WebGL surface does not need every cell. For the browser we
sample the matrix to at most `interactive_max_dim` cells on its longest side;
this keeps the HTML widget small enough to render and rotate smoothly.

Because both the projected DEM spacing and elevation values are measured in
metres, the natural `zscale` follows the effective grid resolution. Dividing it
by `vertical_exaggeration` makes the relief easier to read without changing the
underlying elevation data.

The circuit itself stays as `sf`: `render_path()` accepts spatial line geometry
and positions it against the same projected extent.

```{r}
#| label: render-interactive-terrain
#| column: page
#| fig-alt: "Interactive 3D terrain around the Nürburgring with the circuit traced in a pale line over green relief."

library(rayshader)

# Reduce only the interactive mesh; keep the full-resolution elevation in the RDS.
interactive_step <- max(1L, ceiling(max(dim(elevation)) / interactive_max_dim))
interactive_rows <- seq.int(1L, nrow(elevation), by = interactive_step)
interactive_cols <- seq.int(1L, ncol(elevation), by = interactive_step)
elevation_interactive <- elevation[
  interactive_rows,
  interactive_cols,
  drop = FALSE
]
interactive_resolution_m <- resolution_m * interactive_step

zscale <- interactive_resolution_m / vertical_exaggeration

terrain_palette <- grDevices::colorRampPalette(
  c("#203529", forest_green, "#94AD91")
)(256)

terrain_texture <- elevation_interactive |>
  rayshader::height_shade(texture = terrain_palette) |>
  rayshader::add_shadow(
    rayshader::lamb_shade(
      elevation_interactive,
      zscale = zscale,
      sunangle = 315,
      sunaltitude = 45
    ),
    max_darken = 0.35
  )

terrain_texture |>
  rayshader::plot_3d(
    elevation_interactive,
    zscale = zscale,
    theta = 145,
    phi = 35,
    zoom = 0.72,
    fov = 35,
    windowsize = c(1200, 675),
    background = "#F7F8F8",
    shadow = FALSE
  )

rayshader::render_path(
  lat = track,
  extent = terrain_extent,
  heightmap = elevation_interactive,
  zscale = zscale,
  linewidth = 5,
  color = track_colour,
  antialias = TRUE,
  offset = track_offset_m
)

rgl::rglwidget(
  width = 1200,
  height = 675
)
```

This is intentionally only the first pass: terrain plus circuit. Once the scene
and WebGL export behave well, we can decide whether roads, labels, forest texture
or a high-quality static render actually improve it.

Data: [© OpenStreetMap contributors](https://www.openstreetmap.org/copyright),
licensed under the ODbL. Elevation is retrieved with
[`elevatr`](https://cran.r-project.org/package=elevatr) from AWS terrain tiles.
