---
title: Nürburgring in 3D with rayshader
description: Trace the Nürburgring over real elevation and explore how camera angle changes the circuit in 3D.
author: Joshua Kunst
date: 2026-09-10
categories: [R, maps, rayshader, OpenStreetMap]
lang: en
draft: true
toc: true
---

```{r}
#| label: setup
#| include: false

source("../../_R/post_setup.R")

install_missing_packages(c(
  "dplyr",
  "elevatr",
  "osmdata",
  "rayshader",
  "rgl",
  "sf",
  "terra"
))

# Render WebGL without requiring a visible OpenGL window during Quarto rendering.
options(rgl.useNULL = TRUE, rgl.printRglwidget = FALSE)
rgl::setupKnitr(autoprint = FALSE)
```

The [2D Nürburgring post](../2026-09-10-nurburgring-one-geom-at-a-time/)
builds the circuit from OpenStreetMap geometries with `sf` and `ggplot2`. Here we
keep that part deliberately short and focus on one new dimension: elevation.

This version is intentionally narrower. The circuit defines the frame, the terrain
provides the shape underneath it, and the camera does the rest. No surrounding
road network is needed.

## Track area

We still start near the Nürburgring to find its OpenStreetMap raceway geometries,
but the final terrain frame is derived from the track itself. This keeps the 3D
surface compact and makes the circuit fill more of the screen.

```{r}
#| label: terrain-parameters

library(dplyr)
library(sf)

centre_lon <- 6.95
centre_lat <- 50.35
search_width_km <- 16
screen_ratio <- 16 / 9
terrain_margin_pct <- 10

forest_green <- "#47714D"
asphalt_dark <- "#30373D"
track_colour <- "#EAF1E8"
vertical_exaggeration <- 2
track_offset_m <- 2
interactive_max_dim <- 400

# Keep camera settings together so different captures are easy to reproduce.
camera_theta <- 145
camera_phi <- 42
camera_zoom <- 0.78
camera_fov <- 35

centre_utm <- sf::st_sfc(
  sf::st_point(c(centre_lon, centre_lat)),
  crs = 4326
) |>
  sf::st_transform(25832)

centre_xy <- sf::st_coordinates(centre_utm)[1, ]
search_half_m <- search_width_km * 1000 / 2

# This broad square is used only to find the circuit; it is not the final terrain frame.
search_frame_utm <- sf::st_bbox(
  c(
    xmin = centre_xy[[1]] - search_half_m,
    ymin = centre_xy[[2]] - search_half_m,
    xmax = centre_xy[[1]] + search_half_m,
    ymax = centre_xy[[2]] + search_half_m
  ),
  crs = sf::st_crs(25832)
) |>
  sf::st_as_sfc()

search_bbox_wgs84 <- search_frame_utm |>
  sf::st_transform(4326) |>
  sf::st_bbox()
```

## Data

The post keeps a single processed RDS. On the first run we find the raceway ways,
use their bounds to request only the terrain around the circuit, and store the
processed track and elevation together. Subsequent renders use that local bundle.

```{r}
#| label: prepare-terrain-data
#| message: false

cache_file <- "data/nurburgring-3d.rds"
cache_version <- 2L

terrain_data <- if (file.exists(cache_file)) readRDS(cache_file) else NULL
cache_is_current <- !is.null(terrain_data$cache_version) && terrain_data$cache_version == cache_version

if (!cache_is_current) {
  raceway_query <- osmdata::opq(
    bbox = search_bbox_wgs84,
    timeout = 180
  ) |>
    osmdata::add_osm_feature(
      key = "highway",
      value = "raceway"
    )

  track <- osmdata::osmdata_sf(raceway_query)$osm_lines |>
    filter(highway == "raceway") |>
    select(osm_id, name, geometry) |>
    sf::st_transform(25832) |>
    sf::st_intersection(search_frame_utm)

  if (nrow(track) == 0) {
    stop("No raceway geometry found around the Nürburgring search area.")
  }

  # Start from the circuit bounds, then add a small margin around all raceway geometry.
  track_bbox <- sf::st_bbox(track)
  track_width_m <- track_bbox[["xmax"]] - track_bbox[["xmin"]]
  track_height_m <- track_bbox[["ymax"]] - track_bbox[["ymin"]]
  track_centre <- c(
    (track_bbox[["xmin"]] + track_bbox[["xmax"]]) / 2,
    (track_bbox[["ymin"]] + track_bbox[["ymax"]]) / 2
  )

  frame_width_m <- track_width_m * (1 + terrain_margin_pct / 100)
  frame_height_m <- track_height_m * (1 + terrain_margin_pct / 100)

  # Expand only the shorter side so the terrain itself follows the 16:9 capture ratio.
  if (frame_width_m / frame_height_m < screen_ratio) {
    frame_width_m <- frame_height_m * screen_ratio
  } else {
    frame_height_m <- frame_width_m / screen_ratio
  }

  terrain_frame_utm <- sf::st_bbox(
    c(
      xmin = track_centre[[1]] - frame_width_m / 2,
      ymin = track_centre[[2]] - frame_height_m / 2,
      xmax = track_centre[[1]] + frame_width_m / 2,
      ymax = track_centre[[2]] + frame_height_m / 2
    ),
    crs = sf::st_crs(25832)
  ) |>
    sf::st_as_sfc()

  terrain_bbox_wgs84 <- terrain_frame_utm |>
    sf::st_transform(4326) |>
    sf::st_bbox()

  # Two corner points are enough for elevatr to determine the required AWS tiles.
  elev_locations <- data.frame(
    x = c(terrain_bbox_wgs84[["xmin"]], terrain_bbox_wgs84[["xmax"]]),
    y = c(terrain_bbox_wgs84[["ymin"]], terrain_bbox_wgs84[["ymax"]])
  )

  dem <- elevatr::get_elev_raster(
    locations = elev_locations,
    z = 13,
    prj = "EPSG:4326",
    src = "aws",
    verbose = TRUE
  )

  # elevatr 0.99.x returns RasterLayer; upcoming versions return SpatRaster.
  if (!inherits(dem, "SpatRaster")) {
    dem <- methods::as(dem, "SpatRaster")
  }

  dem <- dem |>
    terra::project("EPSG:25832", method = "bilinear")

  # Crop after projection so the stored terrain follows the circuit-derived frame exactly.
  terrain_extent_utm <- terra::ext(terra::vect(terrain_frame_utm))
  dem <- terra::crop(dem, terrain_extent_utm)

  elevation <- rayshader::raster_to_matrix(dem)
  terrain_extent <- as.vector(terra::ext(dem))
  names(terrain_extent) <- c("xmin", "xmax", "ymin", "ymax")
  resolution_m <- mean(terra::res(dem))

  dir.create("data", recursive = TRUE, showWarnings = FALSE)
  saveRDS(
    list(
      cache_version = cache_version,
      track = track,
      elevation = elevation,
      extent = terrain_extent,
      resolution_m = resolution_m
    ),
    cache_file
  )

  terrain_data <- readRDS(cache_file)
}

track <- terrain_data$track
elevation <- terrain_data$elevation
terrain_extent <- terrain_data$extent
resolution_m <- terrain_data$resolution_m

c(
  elevation_min_m = min(elevation, na.rm = TRUE),
  elevation_max_m = max(elevation, na.rm = TRUE),
  grid_resolution_m = resolution_m
)
```

## Nürburgring in 3D

The cached DEM keeps its downloaded detail, but an interactive WebGL surface does
not need every cell. We sample the matrix only for the browser scene, keeping the
RDS at full resolution for later high-quality captures.

The terrain texture stays abstract rather than photographic: elevation drives a
small family of forest greens, while directional shading reveals the folds in the
land. A little vertical exaggeration makes those changes easier to read.

```{r}
#| label: render-interactive-terrain
#| column: page
#| fig-alt: "Interactive 3D terrain tightly framed around the Nürburgring, with the circuit traced over green relief."

library(rayshader)

# Reduce only the interactive mesh; keep the full-resolution elevation in the RDS.
interactive_step <- max(1L, ceiling(max(dim(elevation)) / interactive_max_dim))
interactive_rows <- seq.int(1L, nrow(elevation), by = interactive_step)
interactive_cols <- seq.int(1L, ncol(elevation), by = interactive_step)
elevation_interactive <- elevation[
  interactive_rows,
  interactive_cols,
  drop = FALSE
]
interactive_resolution_m <- resolution_m * interactive_step
zscale <- interactive_resolution_m / vertical_exaggeration

# Use several nearby greens so height variation is visible without resembling satellite imagery.
terrain_palette <- grDevices::colorRampPalette(
  c("#18281F", "#2D4933", forest_green, "#718E70", "#A7BAA1")
)(256)

terrain_texture <- elevation_interactive |>
  rayshader::height_shade(texture = terrain_palette) |>
  rayshader::add_shadow(
    rayshader::lamb_shade(
      elevation_interactive,
      zscale = zscale,
      sunangle = 315,
      sunaltitude = 40
    ),
    max_darken = 0.42
  )

terrain_texture |>
  rayshader::plot_3d(
    elevation_interactive,
    zscale = zscale,
    theta = camera_theta,
    phi = camera_phi,
    zoom = camera_zoom,
    fov = camera_fov,
    windowsize = c(1200, 675),
    background = "#F7F8F8",
    shadow = FALSE,
    solid = FALSE
  )

# A dark casing separates the pale circuit from both bright and shaded terrain.
rayshader::render_path(
  lat = track,
  extent = terrain_extent,
  heightmap = elevation_interactive,
  zscale = zscale,
  linewidth = 7,
  color = asphalt_dark,
  antialias = TRUE,
  offset = track_offset_m
)

rayshader::render_path(
  lat = track,
  extent = terrain_extent,
  heightmap = elevation_interactive,
  zscale = zscale,
  linewidth = 4,
  color = track_colour,
  antialias = TRUE,
  offset = track_offset_m + 0.5
)

rgl::rglwidget(
  width = 1200,
  height = 675
)
```

## Camera first

The interactive scene is useful for finding viewpoints before spending time on a
high-quality render. `camera_theta`, `camera_phi`, `camera_zoom` and `camera_fov`
are kept with the other parameters so the same view can later be reproduced as a
snapshot or ray-traced image.

For now the experiment stops here: one circuit, one terrain surface and one camera.
The next step is to explore a few deliberate viewpoints rather than add more map
layers.

Data: [© OpenStreetMap contributors](https://www.openstreetmap.org/copyright),
licensed under the ODbL. Elevation is retrieved with
[`elevatr`](https://cran.r-project.org/package=elevatr) from AWS terrain tiles.
